---
title: "Pakistan Inflation Dashboard"
subtitle: "Underlying Inflation Measures & Monetary Policy Tracker"
author: "State Bank of Pakistan Analysis"
date: "September 2025"
format:
  dashboard:
    document-css: false
    link-citations: true
    date-format: long
    lang: en
    page-layout: custom
    theme: cosmo
    # logo: ""  # Remove this line or comment it out
    nav-buttons:
      - icon: github
        href: https://github.com
    orientation: columns
---

```{r}
#| label: setup
#| include: false

# Load required libraries
library(plotly)
library(dplyr)
library(tidyr)
library(knitr)
library(DT)
library(htmltools)
library(kableExtra)

# Current inflation measures (September 2025)
inflation_measures <- data.frame(
  name = c("Headline CPI", "Core CPI", "Food Inflation", 
           "Non-Food Inflation", "Urban CPI", "Rural CPI"),
  current = c(5.6, 7.0, 3.8, 8.2, 5.8, 5.3),
  previous = c(3.0, 7.2, 0.5, 8.5, 3.2, 2.7),
  target = c(6.0, 6.5, 6.0, 6.0, 6.0, 6.0),
  description = c("Overall consumer price inflation",
                  "Excludes food and energy",
                  "Food and beverages component",
                  "Excluding food items",
                  "Urban areas inflation",
                  "Rural areas inflation"),
  stringsAsFactors = FALSE
)

# Add calculated columns
inflation_measures <- inflation_measures %>%
  mutate(
    change = current - previous,
    diff_from_target = current - target,
    status = case_when(
      abs(diff_from_target) <= 0.5 ~ "Within Target",
      diff_from_target < -0.5 ~ "Below Target",
      diff_from_target > 0.5 & diff_from_target <= 2 ~ "Above Target",
      TRUE ~ "Significantly Above"
    )
  )

# Historical data (last 24 months)
historical_data <- data.frame(
  month = c("Oct-23", "Nov-23", "Dec-23", "Jan-24", "Feb-24", "Mar-24",
            "Apr-24", "May-24", "Jun-24", "Jul-24", "Aug-24", "Sep-24",
            "Oct-24", "Nov-24", "Dec-24", "Jan-25", "Feb-25", "Mar-25",
            "Apr-25", "May-25", "Jun-25", "Jul-25", "Aug-25", "Sep-25"),
  headline = c(26.9, 29.2, 29.7, 28.3, 23.1, 20.7, 17.3, 11.8, 12.6, 
               11.1, 9.6, 6.9, 7.2, 4.9, 4.1, 2.7, 1.8, 0.7, 0.5, 
               3.5, 3.2, 4.5, 3.0, 5.6),
  core = c(18.5, 19.2, 20.1, 19.8, 17.5, 16.2, 15.1, 14.2, 13.8, 
           12.5, 11.3, 10.2, 9.8, 9.1, 8.5, 8.2, 7.9, 7.6, 7.4, 
           7.3, 7.1, 7.0, 7.2, 7.0),
  food = c(30.5, 34.1, 35.2, 32.8, 26.5, 23.8, 18.5, 10.2, 11.5, 
           9.8, 7.5, 3.8, 4.2, 1.5, 0.8, -1.2, -2.5, -4.1, -4.8, 
           1.2, 0.8, 2.5, -0.5, 3.8),
  policy = c(22.0, 22.0, 22.0, 22.0, 22.0, 22.0, 22.0, 22.0, 22.0,
             19.5, 19.5, 17.5, 17.5, 15.0, 15.0, 12.0, 12.0, 12.0,
             12.0, 11.0, 11.0, 11.0, 11.0, 11.0),
  stringsAsFactors = FALSE
)

# Component data
component_data <- data.frame(
  category = c("Food & Beverages", "Food & Beverages", "Food & Beverages",
               "Housing & Utilities", "Housing & Utilities", "Housing & Utilities",
               "Transportation", "Transportation",
               "Other", "Other", "Other"),
  item = c("Perishable Foods", "Non-Perishable Foods", "Beverages",
           "Electricity", "Gas", "Housing Rent",
           "Motor Fuel", "Transport Services",
           "Healthcare", "Education", "Communication"),
  value = c(-2.5, 8.2, 12.5, 25.8, 35.2, 4.5, -8.5, 15.3, 9.8, 14.2, 3.1),
  stringsAsFactors = FALSE
)

# Key statistics
current_policy_rate <- 11.0
current_headline <- 5.6
real_rate <- current_policy_rate - current_headline
inflation_expectations <- 66.6
```

# Overview {orientation="columns"}

## Column {width=25%}

```{r}
#| content: valuebox
#| title: "Current Policy Rate"

list(
  value = paste0(current_policy_rate, "%"),
  caption = "Unchanged since May 2025",
  color = "primary"
)
```

```{r}
#| content: valuebox
#| title: "Real Interest Rate"

list(
  value = paste0(round(real_rate, 1), "%"),
  caption = "Policy Rate - Headline CPI",
  color = "success"
)
```

```{r}
#| content: valuebox
#| title: "Inflation Expectations"

list(
  value = round(inflation_expectations, 1),
  caption = "Consumer Expectations Index ↓ 0.6",
  color = "info"
)
```

## Column {width=75%}

### Inflation Trends & Monetary Policy Rate

```{r}
#| title: "24-Month Historical View"
#| padding: 0

# Create ordered factor for proper month ordering
historical_data_ordered <- historical_data %>%
  mutate(month_ordered = factor(month, levels = month))

plot_ly(historical_data_ordered) %>%
  add_trace(
    x = ~month_ordered, 
    y = ~headline, 
    type = 'scatter', 
    mode = 'lines+markers',
    name = 'Headline CPI', 
    line = list(color = '#3b82f6', width = 3),
    marker = list(size = 4)
  ) %>%
  add_trace(
    x = ~month_ordered, 
    y = ~core, 
    type = 'scatter', 
    mode = 'lines+markers',
    name = 'Core CPI', 
    line = list(color = '#f59e0b', width = 2),
    marker = list(size = 4)
  ) %>%
  add_trace(
    x = ~month_ordered, 
    y = ~food, 
    type = 'scatter', 
    mode = 'lines+markers',
    name = 'Food Inflation', 
    line = list(color = '#10b981', width = 2),
    marker = list(size = 4)
  ) %>%
  add_trace(
    x = ~month_ordered, 
    y = ~policy, 
    type = 'scatter', 
    mode = 'lines',
    name = 'Policy Rate', 
    line = list(color = '#ef4444', width = 3, dash = 'dash')
  ) %>%
  layout(
    xaxis = list(
      title = "", 
      tickangle = -45,
      categoryorder = "trace",
      tickfont = list(size = 10)
    ),
    yaxis = list(
      title = "Percentage (%)",
      titlefont = list(size = 12)
    ),
    hovermode = 'x unified',
    legend = list(
      x = 0.02, 
      y = 0.98,
      bgcolor = 'rgba(255, 255, 255, 0.8)',
      bordercolor = '#e2e8f0',
      borderwidth = 1
    ),
    margin = list(l = 50, r = 20, t = 40, b = 80)
  )
```

::: {.callout-note}
## Key Insight
Pakistan's policy rate has been cut from 22% to 11% since June 2024, as inflation dropped from a peak of nearly 30% to current levels. The State Bank maintained rates at 11% since May 2025 due to energy price adjustments and flood-related risks.
:::

# Inflation Measures {orientation="rows"}

## Row {height=60%}

### Current Inflation Measures (September 2025)

```{r}
#| title: "Inflation Indicators - Status Relative to Target"

# Create formatted table
table_data <- inflation_measures %>%
  mutate(
    Measure = name,
    Description = description,
    Current = paste0(sprintf("%.1f", current), "%"),
    Change = paste0(ifelse(change > 0, "+", ""), sprintf("%.1f", change), "pp"),
    Target = paste0(sprintf("%.1f", target), "%"),
    Status = status
  ) %>%
  select(Measure, Description, Current, Change, Target, Status)

# Color mapping
color_map <- c(
  "Within Target" = "#d4edda",
  "Below Target" = "#d1ecf1",
  "Above Target" = "#fff3cd",
  "Significantly Above" = "#f8d7da"
)

datatable(
  table_data,
  options = list(
    dom = 't',
    pageLength = 10,
    ordering = FALSE,
    columnDefs = list(
      list(width = '200px', targets = 0),
      list(width = '300px', targets = 1),
      list(className = 'dt-center', targets = 2:5)
    )
  ),
  rownames = FALSE,
  class = 'cell-border stripe'
) %>%
  formatStyle(
    'Status',
    backgroundColor = styleEqual(
      names(color_map),
      unname(color_map)
    ),
    fontWeight = 'bold'
  ) %>%
  formatStyle(
    'Current',
    fontWeight = 'bold',
    fontSize = '16px'
  )
```

## Row {height=40%}

### Target Range Explanation

**Understanding the Status Colors:**

- **🟢 Within Target (Green):** Inflation is within ±0.5 percentage points of the target, indicating price stability.

- **🔵 Below Target (Blue):** Inflation is more than 0.5pp below target, which may indicate weak demand or deflationary pressures.

- **🟡 Above Target (Yellow):** Inflation is 0.5-2.0pp above target, suggesting moderate price pressures that require monitoring.

- **🔴 Significantly Above (Red):** Inflation is more than 2.0pp above target, indicating substantial price pressures requiring policy action.

The State Bank of Pakistan targets inflation in the range of **5-7%** to maintain price stability while supporting economic growth.

# Components {orientation="rows"}

## Row {height=60%}

### Inflation by Component Category

```{r}
#| title: "Year-over-Year Changes by Category"

# Reorder items by value for better visualization
component_data_sorted <- component_data %>%
  arrange(value) %>%
  mutate(
    item_ordered = factor(item, levels = item),
    color = case_when(
      value > 10 ~ "#dc3545",
      value > 0 ~ "#ffc107",
      TRUE ~ "#28a745"
    )
  )

plot_ly(component_data_sorted) %>%
  add_trace(
    y = ~item_ordered,
    x = ~value,
    type = 'bar',
    orientation = 'h',
    marker = list(color = ~color),
    text = ~paste0(sprintf("%.1f", value), "%"),
    textposition = 'outside',
    textfont = list(size = 11),
    hovertemplate = paste(
      '<b>%{y}</b><br>',
      'Change: %{x:.1f}%<br>',
      '<extra></extra>'
    )
  ) %>%
  layout(
    xaxis = list(
      title = "Year-over-Year Change (%)",
      titlefont = list(size = 12)
    ),
    yaxis = list(
      title = "",
      tickfont = list(size = 10)
    ),
    showlegend = FALSE,
    margin = list(l = 150, r = 50, t = 40, b = 50)
  )
```

## Row {height=40%}

### Food & Beverages

```{r}
food_items <- component_data %>% 
  filter(category == "Food & Beverages")

kable(
  food_items %>% select(Item = item, `YoY Change (%)` = value),
  align = c('l', 'r'),
  format = 'html'
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

### Housing & Utilities

```{r}
housing_items <- component_data %>% 
  filter(category == "Housing & Utilities")

kable(
  housing_items %>% select(Item = item, `YoY Change (%)` = value),
  align = c('l', 'r'),
  format = 'html'
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

### Transportation

```{r}
transport_items <- component_data %>% 
  filter(category == "Transportation")

kable(
  transport_items %>% select(Item = item, `YoY Change (%)` = value),
  align = c('l', 'r'),
  format = 'html'
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

### Other Categories

```{r}
other_items <- component_data %>% 
  filter(category == "Other")

kable(
  other_items %>% select(Item = item, `YoY Change (%)` = value),
  align = c('l', 'r'),
  format = 'html'
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

# Time Series Analysis {orientation="rows"}

## Row {height=50%}

### Headline vs Core Inflation

```{r}
# Create ordered factor for proper chronological display
historical_data_ordered <- historical_data %>%
  mutate(month_ordered = factor(month, levels = month))

plot_ly(historical_data_ordered) %>%
  add_trace(
    x = ~month_ordered, 
    y = ~headline, 
    type = 'scatter', 
    mode = 'lines+markers',
    name = 'Headline CPI', 
    line = list(color = '#3b82f6', width = 3),
    marker = list(size = 5)
  ) %>%
  add_trace(
    x = ~month_ordered, 
    y = ~core, 
    type = 'scatter', 
    mode = 'lines+markers',
    name = 'Core CPI', 
    line = list(color = '#f59e0b', width = 3),
    marker = list(size = 5)
  ) %>%
  add_trace(
    x = ~month_ordered, 
    y = 6.5,
    type = 'scatter',
    mode = 'lines',
    name = 'Target (6.5%)',
    line = list(color = 'gray', width = 2, dash = 'dot')
  ) %>%
  layout(
    xaxis = list(
      title = "", 
      tickangle = -45,
      categoryorder = "trace",
      tickfont = list(size = 10)
    ),
    yaxis = list(
      title = "Percentage (%)",
      titlefont = list(size = 12)
    ),
    hovermode = 'x unified',
    legend = list(
      x = 0.02, 
      y = 0.98,
      bgcolor = 'rgba(255, 255, 255, 0.8)',
      bordercolor = '#e2e8f0',
      borderwidth = 1
    ),
    margin = list(l = 50, r = 20, t = 20, b = 80)
  )
```

### Food Inflation Trend

```{r}
# Create ordered factor
historical_data_ordered <- historical_data %>%
  mutate(month_ordered = factor(month, levels = month))

plot_ly(historical_data_ordered) %>%
  add_trace(
    x = ~month_ordered, 
    y = ~food, 
    type = 'scatter', 
    mode = 'lines',
    name = 'Food Inflation',
    fill = 'tozeroy',
    fillcolor = 'rgba(16, 185, 129, 0.2)',
    line = list(color = '#10b981', width = 3)
  ) %>%
  layout(
    xaxis = list(
      title = "", 
      tickangle = -45,
      categoryorder = "trace",
      tickfont = list(size = 10)
    ),
    yaxis = list(
      title = "Percentage (%)",
      titlefont = list(size = 12)
    ),
    hovermode = 'x',
    margin = list(l = 50, r = 20, t = 20, b = 80)
  )
```

## Row {height=50%}

### Policy Rate Evolution

```{r}
# Create ordered factor
historical_data_ordered <- historical_data %>%
  mutate(month_ordered = factor(month, levels = month))

plot_ly(historical_data_ordered) %>%
  add_trace(
    x = ~month_ordered, 
    y = ~policy, 
    type = 'scatter', 
    mode = 'lines+markers',
    name = 'Policy Rate',
    line = list(color = '#ef4444', width = 4),
    marker = list(size = 7, symbol = 'diamond')
  ) %>%
  layout(
    xaxis = list(
      title = "", 
      tickangle = -45,
      categoryorder = "trace",
      tickfont = list(size = 10)
    ),
    yaxis = list(
      title = "Policy Rate (%)",
      titlefont = list(size = 12)
    ),
    hovermode = 'x',
    margin = list(l = 50, r = 20, t = 20, b = 80)
  )
```

### Inflation Differential

```{r}
# Create ordered factor and calculate differential
historical_data_diff <- historical_data %>%
  mutate(
    month_ordered = factor(month, levels = month),
    diff = headline - core
  )

plot_ly(historical_data_diff) %>%
  add_trace(
    x = ~month_ordered, 
    y = ~diff, 
    type = 'bar',
    name = 'Headline - Core Differential',
    marker = list(
      color = ~diff,
      colorscale = list(
        c(0, 'rgb(239, 68, 68)'),   # Red for negative
        c(0.5, 'rgb(255, 255, 255)'),  # White for zero
        c(1, 'rgb(34, 197, 94)')    # Green for positive
      ),
      showscale = FALSE
    ),
    hovertemplate = paste(
      '<b>%{x}</b><br>',
      'Differential: %{y:.2f}pp',
      '<extra></extra>'
    )
  ) %>%
  layout(
    xaxis = list(
      title = "", 
      tickangle = -45,
      categoryorder = "trace",
      tickfont = list(size = 10)
    ),
    yaxis = list(
      title = "Differential (pp)",
      titlefont = list(size = 12)
    ),
    hovermode = 'x',
    margin = list(l = 50, r = 20, t = 20, b = 80)
  )
```

# About {orientation="rows"}

## Row

### About This Dashboard

#### Purpose
This dashboard tracks multiple measures of inflation in Pakistan to provide a comprehensive view of price pressures in the economy, similar to the Federal Reserve Bank of Atlanta's Underlying Inflation Dashboard.

#### Data Sources
- **State Bank of Pakistan (SBP)** - Monetary policy and inflation data
- **Pakistan Bureau of Statistics (PBS)** - Consumer Price Index
- **International Monetary Fund (IMF)** - Economic indicators

#### Update Frequency
**Monthly** - Data is updated following the release of official CPI figures, typically in the first half of each month.

#### Methodology
Inflation measures are color-coded based on their distance from the SBP's target range of 5-7%:

- **🟢 Green (Within Target):** ±0.5pp from target
- **🟡 Yellow (Above Target):** 0.5-2.0pp above target  
- **🔴 Red (Significantly Above):** >2.0pp above target
- **🔵 Blue (Below Target):** >0.5pp below target

### Key Indicators

#### Headline CPI
Overall consumer price inflation measuring the average change in prices paid by consumers for a basket of goods and services.

#### Core CPI
Excludes volatile food and energy prices to show underlying inflation trends. This measure is closely watched by the State Bank of Pakistan for monetary policy decisions.

#### Food Inflation
Measures price changes in food and beverages, which constitute a significant portion of household expenditure in Pakistan.

#### Policy Rate
The State Bank of Pakistan's benchmark interest rate, used to control inflation and influence economic activity. Current rate: **11.0%**

#### Real Interest Rate
The policy rate adjusted for inflation (Policy Rate - Headline CPI). Positive real rates help control inflation by encouraging savings over spending.

### Recent Developments

#### Monetary Policy Actions (2024-2025)

- **June 2024:** Started rate cutting cycle from 22%
- **Multiple cuts:** Reduced rates by 1,100 basis points total
- **May 2025:** Final cut to 11%
- **June-September 2025:** Held rates steady at 11%

#### Key Factors

**Disinflation Success:**
- Inflation fell from 30% peak to current 5.6%
- Food prices moderated significantly
- Core inflation remains elevated at 7%

**Ongoing Challenges:**
- Energy tariff adjustments (gas +35%, electricity +26%)
- Flood-related agricultural impacts
- Global commodity price volatility
- Exchange rate pressures

**Outlook:**
The State Bank expects inflation to stabilize in the 5-7% target range, with continued focus on maintaining positive real interest rates to anchor inflation expectations.

### Contact & Feedback

For questions, suggestions, or feedback about this dashboard:

📧 Email: analytics@example.com  
🌐 Website: www.sbp.org.pk  
📱 Twitter: @StateBank_Pak

---

**Dashboard Version:** 1.0  
**Last Updated:** September 2025  
**Next Update:** Mid-October 2025

*This dashboard is for informational purposes only and does not constitute financial advice or policy guidance from the State Bank of Pakistan.*