---
title: "ARDL and ARDL Cointegration"
subtitle: "A Practical Guide with R"
author: "Zahid Asghar"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  revealjs:
    theme: [default, ../assets/slides.scss]
    chalkboard: true
    slide-number: c/t
    preview-links: auto
    incremental: false
    code-fold: false
    title-slide-attributes:
      data-background-color: "#cf3a36"
execute: 
  freeze: auto
  echo: true
  warning: false
  message: false
---

## Outline

::: {.incremental}
1. Review: Why Cointegration Matters
2. ARDL Models: The Basics
3. ARDL Bounds Testing Approach
4. Step-by-Step Implementation in R
5. Interpreting Results
6. Advantages over Johansen Method
7. Practical Example with Real Data
:::

# Review: Cointegration Basics

## What is Cointegration? {.smaller}

**Recall from Johansen Method:**

- Two or more non-stationary series (I(1)) can have a **stationary linear combination**
- This implies a long-run equilibrium relationship
- Short-run deviations are temporary

**Example:**
$$Y_t = \beta_0 + \beta_1 X_t + u_t$$

If $Y_t \sim I(1)$ and $X_t \sim I(1)$, but $u_t \sim I(0)$, then $Y$ and $X$ are cointegrated.

::: {.callout-note}
Cointegration prevents **spurious regression** problems!
:::

## Johansen Method: Key Requirements {.smaller}

You already know Johansen cointegration requires:

::: {.incremental}
1. **All variables must be I(1)** - same order of integration
2. **System of equations** - VAR framework
3. **Pre-testing for unit roots** is essential
4. **Trace test** or **Maximum eigenvalue test**
5. Assumes symmetric adjustment to equilibrium
:::

::: {.callout-important}
**Question:** What if some variables are I(0) and others are I(1)?  
**Answer:** This is where ARDL comes in!
:::

# ARDL Models

## What is ARDL? {.smaller}

**AutoRegressive Distributed Lag Model**

General form with one dependent and one independent variable:

$$Y_t = \alpha_0 + \sum_{i=1}^{p}\beta_i Y_{t-i} + \sum_{j=0}^{q}\gamma_j X_{t-j} + \varepsilon_t$$

::: {.incremental}
- **AutoRegressive (AR):** lags of dependent variable $Y_{t-1}, Y_{t-2}, ...$
- **Distributed Lag (DL):** current and lagged values of $X_t, X_{t-1}, X_{t-2}, ...$
- **p:** lag order for Y (AR part)
- **q:** lag order for X (DL part)
- Denoted as **ARDL(p, q)**
:::

## Why Use ARDL? {.smaller}

**Flexibility in modeling dynamic relationships:**

::: {.incremental}
1. Captures **short-run dynamics** (through differenced terms)
2. Estimates **long-run relationships** (through level terms)
3. Allows **different lag lengths** for different variables
4. Can include **exogenous variables** easily
5. Works with **small sample sizes** (minimum ~30 observations)
:::

::: {.callout-tip}
ARDL is particularly useful for macroeconomic data where relationships have both immediate and lagged effects.
:::

## ARDL with Multiple Variables {.smaller}

With multiple independent variables:

$$Y_t = \alpha_0 + \sum_{i=1}^{p}\beta_i Y_{t-i} + \sum_{j=0}^{q_1}\gamma_{1j} X_{1,t-j} + \sum_{j=0}^{q_2}\gamma_{2j} X_{2,t-j} + \sum_{j=0}^{q_3}\gamma_{3j} X_{3,t-j} + \varepsilon_t$$

::: {.incremental}
- Denoted as **ARDL(p, q₁, q₂, q₃)**
- Each variable can have **different optimal lag length**
- Example: ARDL(2, 1, 3, 0) means:
  - 2 lags of Y
  - 1 lag of X₁
  - 3 lags of X₂
  - 0 lags (contemporaneous only) of X₃
:::

# ARDL Bounds Testing Approach

## The Innovation: Pesaran, Shin & Smith (2001) {.smaller}

**The PSS (2001) Bounds Testing Approach**

::: {.incremental}
- Revolutionary paper: "Bounds testing approaches to the analysis of level relationships"
- Key innovation: Test for cointegration **without pre-testing** for unit roots
- Works with variables that are:
  - I(0) - stationary
  - I(1) - first difference stationary
  - **Mix of I(0) and I(1)**
:::

::: {.callout-warning}
**Important restriction:** Variables cannot be I(2) or higher!  
You still need to check this.
:::

## How Does It Work? {.smaller}

**Two-step transformation:**

**Step 1:** Original ARDL model in levels

**Step 2:** Transform to **Error Correction Model (ECM)** representation

$$\Delta Y_t = \beta_0 + \sum_{i=1}^{p-1}\beta_i \Delta Y_{t-i} + \sum_{j=0}^{q-1}\gamma_j \Delta X_{t-j} + \underbrace{\theta_0 Y_{t-1} + \theta_1 X_{t-1}}_{\text{Long-run relationship}} + \varepsilon_t$$

::: {.callout-note}
The **lagged level terms** ($Y_{t-1}, X_{t-1}$) capture the long-run relationship!
:::

## The Bounds Test {.smaller}

**Hypothesis Testing:**

$$H_0: \theta_0 = \theta_1 = 0 \quad \text{(No cointegration)}$$
$$H_1: \theta_0 \neq 0 \text{ or } \theta_1 \neq 0 \quad \text{(Cointegration exists)}$$

Test using **F-statistic** (or Wald test)

::: {.incremental}
- **But:** F-statistic has non-standard distribution
- **Solution:** PSS provide **critical value bounds**
  - **Lower bound I(0):** assumes all variables are I(0)
  - **Upper bound I(1):** assumes all variables are I(1)
:::

## Decision Rules {.smaller}

Compare your **F-statistic** with critical value bounds:

```{r}
#| echo: false
#| fig-height: 3

library(ggplot2)

df <- data.frame(
  x = c(0, 5, 10, 15),
  y = c(0, 0, 0, 0),
  label = c("", "I(0) Lower", "I(1) Upper", "")
)

ggplot(df, aes(x = x, y = y)) +
  geom_segment(aes(x = 5, xend = 5, y = -0.2, yend = 0.2), 
               color = "blue", size = 1.5) +
  geom_segment(aes(x = 10, xend = 10, y = -0.2, yend = 0.2), 
               color = "red", size = 1.5) +
  annotate("text", x = 2.5, y = 0.3, label = "NO Cointegration", size = 4) +
  annotate("text", x = 7.5, y = 0.3, label = "Inconclusive", size = 4, color = "orange") +
  annotate("text", x = 12.5, y = 0.3, label = "Cointegration!", size = 4, color = "darkgreen") +
  annotate("text", x = 5, y = -0.3, label = "Lower Bound", size = 3, color = "blue") +
  annotate("text", x = 10, y = -0.3, label = "Upper Bound", size = 3, color = "red") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank()) +
  xlim(0, 15) +
  ylim(-0.5, 0.5)
```

::: {.incremental}
1. **F < Lower bound** → No cointegration (reject relationship)
2. **Lower < F < Upper** → Inconclusive (need unit root tests)
3. **F > Upper bound** → Cointegration exists! (proceed to estimate)
:::

## Critical Values {.smaller}

PSS (2001) provide tables based on:

- **Number of variables (k)**
- **Sample size**
- **Significance level** (1%, 5%, 10%)
- **Model specification** (with/without trend)

Example critical values at 5% level (k=3, n=large):

| Bound Type | Critical Value |
|------------|----------------|
| I(0)       | 3.23          |
| I(1)       | 4.35          |

::: {.callout-tip}
Modern R packages calculate these automatically!
:::

# Step-by-Step ARDL Process

## The Complete Workflow {.smaller}

**Step 1: Check Order of Integration**

- Ensure no I(2) variables using ADF/PP tests
- Variables can be I(0), I(1), or mixed

**Step 2: Estimate ARDL Model**

- Select optimal lag lengths using AIC/BIC
- Estimate the unrestricted ARDL model

**Step 3: Bounds Test**

- Test for cointegration using F-statistic
- Compare with critical bounds

## Workflow Continued {.smaller}

**Step 4: If Cointegration Exists**

Estimate the **long-run relationship:**

$$Y_t = \beta_0 + \beta_1 X_{1,t} + \beta_2 X_{2,t} + \beta_3 X_{3,t} + u_t$$

**Step 5: Error Correction Model (ECM)**

$$\Delta Y_t = \text{short-run dynamics} + \delta \cdot ECT_{t-1} + \varepsilon_t$$

Where $ECT_{t-1} = Y_{t-1} - \hat{\beta}_0 - \hat{\beta}_1 X_{1,t-1} - \hat{\beta}_2 X_{2,t-1} - \hat{\beta}_3 X_{3,t-1}$

::: {.callout-important}
$\delta$ should be **negative and significant** for valid cointegration!
:::

# Implementing ARDL in R

## Required Packages {.smaller}

```{r}
#| eval: false

# Install if needed
install.packages("ARDL")
install.packages("dynamac")
install.packages("dplyr")
install.packages("tseries")
install.packages("urca")
install.packages("ggplot2")
```

```{r}
# Load packages
library(ARDL)      # Main ARDL package
library(dynamac)   # Alternative ARDL package
library(dplyr)     # Data manipulation
library(tseries)   # Time series tests
library(urca)      # Unit root tests
library(ggplot2)   # Visualization
```

## Load and Prepare Data {.smaller}

Let's use a practical example: **Relationship between Consumption, Income, and Interest Rate**

```{r}
#| eval: false

# Example with simulated data (replace with your real data)
set.seed(123)
n <- 100

# Create time series
time <- 1:n
income <- 100 + 0.5*time + cumsum(rnorm(n, 0, 2))
interest <- 5 + 0.01*time + rnorm(n, 0, 0.5)
consumption <- 20 + 0.7*income - 2*interest + rnorm(n, 0, 3)

# Create data frame
data <- data.frame(
  time = time,
  consumption = consumption,
  income = income,
  interest = interest
)

head(data)
```

## Step 1: Unit Root Tests {.smaller}

Check that no variables are I(2):

```{r}
#| eval: false

# ADF test function
adf_test <- function(x) {
  test <- ur.df(x, type = "trend", lags = 4)
  return(test@teststat[1])
}

# Test levels
cat("ADF Test - Levels:\n")
cat("Consumption:", adf_test(data$consumption), "\n")
cat("Income:", adf_test(data$income), "\n")
cat("Interest:", adf_test(data$interest), "\n")

# Test first differences
cat("\nADF Test - First Differences:\n")
cat("D.Consumption:", adf_test(diff(data$consumption)), "\n")
cat("D.Income:", adf_test(diff(data$income)), "\n")
cat("D.Interest:", adf_test(diff(data$interest)), "\n")
```

::: {.callout-note}
Critical value at 5%: approximately -3.45 (with trend)
:::

## Step 2: Select Optimal Lags {.smaller}

```{r}
#| eval: false

# Auto-select lags using AIC
ardl_model <- auto_ardl(
  consumption ~ income + interest,
  data = data,
  max_order = 5,    # Maximum lag to consider
  selection = "AIC" # Can also use "BIC"
)

# View selected model
summary(ardl_model)

# Get the selected lag order
ardl_model$order
# Output example: (2, 1, 3) means ARDL(2,1,3)
```

::: {.callout-tip}
**AIC** (Akaike) - better for prediction  
**BIC** (Schwarz) - more parsimonious, better for small samples
:::

## Step 3: Bounds Test {.smaller}

```{r}
#| eval: false

# Perform bounds test for cointegration
bounds_test <- bounds_f_test(
  ardl_model,
  case = 3  # Case 3: unrestricted intercept, no trend
)

# View results
bounds_test

# The output shows:
# - F-statistic
# - Lower bound I(0)
# - Upper bound I(1)
# - Decision
```

**Interpretation:**
```{r}
#| eval: false

if (bounds_test$tab[1,1] > bounds_test$tab[1,3]) {
  cat("F-statistic > Upper bound → Cointegration exists!")
} else if (bounds_test$tab[1,1] < bounds_test$tab[1,2]) {
  cat("F-statistic < Lower bound → No cointegration")
} else {
  cat("Inconclusive → Need to check unit roots")
}
```

## Step 4: Long-Run Relationship {.smaller}

If cointegration is confirmed:

```{r}
#| eval: false

# Extract long-run coefficients
lr_model <- multipliers(ardl_model, type = "lr")

# View long-run relationship
summary(lr_model)

# Long-run elasticities
lr_model$coefficients
```

**Output interpretation:**

- **Coefficient for income:** long-run marginal propensity to consume
- **Coefficient for interest:** long-run effect of interest rate on consumption
- **Standard errors and t-statistics:** test significance of long-run effects

## Step 5: Error Correction Model {.smaller}

```{r}
#| eval: false

# Estimate ECM representation
ecm_model <- ecm(ardl_model)

# View ECM results
summary(ecm_model)

# Extract error correction term coefficient
ect_coef <- ecm_model$coefficients["ect"]

cat("Error Correction Term coefficient:", ect_coef, "\n")
```

**Key checks:**

::: {.incremental}
1. ECT coefficient should be **negative** (adjustment toward equilibrium)
2. ECT coefficient should be **significant** (p < 0.05)
3. Magnitude indicates **speed of adjustment**
   - Example: -0.3 means 30% of disequilibrium corrected each period
:::

## Diagnostic Tests {.smaller}

Always check model validity:

```{r}
#| eval: false

# 1. Serial Correlation Test
bg_test <- bgtest(ardl_model, order = 4)
print(bg_test)

# 2. Heteroskedasticity Test
bp_test <- bptest(ardl_model)
print(bp_test)

# 3. Normality Test
jb_test <- jarque.bera.test(residuals(ardl_model))
print(jb_test)

# 4. Stability Test (CUSUM)
plot(stability(ardl_model, type = "cusum"))
```

::: {.callout-warning}
If diagnostics fail, consider:  
- Adding more lags  
- Including dummy variables  
- Transforming variables (log)
:::

# Interpreting Results

## Understanding Coefficients {.smaller}

**Short-Run Coefficients** (from ARDL/ECM):

```{r}
#| eval: false
#| echo: false

# Example output
#                    Estimate Std.Error t-value Pr(>|t|)
# d(income)           0.450    0.082    5.488   0.000
# d(interest)        -1.234    0.456   -2.706   0.008
```

- $\Delta income$: immediate (impact) effect on consumption
- $\Delta interest$: immediate effect of interest rate change

**Long-Run Coefficients:**

- Show equilibrium relationships
- More policy-relevant for structural analysis

## Error Correction Term {.smaller}

Example: $ECT_{t-1} = -0.35$ (p-value < 0.01)

**Interpretation:**

::: {.incremental}
- **Negative sign**: Confirms cointegration (adjustment toward equilibrium)
- **Magnitude 0.35**: 35% of disequilibrium is corrected each period
- **Time to equilibrium**: Roughly $1/0.35 \approx 2.86$ periods
- **Significance**: The long-run relationship is statistically valid
:::

::: {.callout-important}
If ECT > 0 or insignificant, the cointegration relationship is questionable!
:::

## Visualizing Results {.smaller}

```{r}
#| eval: false

# Plot actual vs fitted values
plot(ardl_model)

# Plot residuals
residuals_df <- data.frame(
  time = 1:length(residuals(ardl_model)),
  residuals = residuals(ardl_model)
)

ggplot(residuals_df, aes(x = time, y = residuals)) +
  geom_line() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "ARDL Model Residuals",
       x = "Time", y = "Residuals") +
  theme_minimal()

# Plot dynamic forecasts
forecast_plot <- forecast(ardl_model, h = 10)
plot(forecast_plot)
```

# ARDL vs Johansen Method

## Comparison Table {.smaller}

| Feature | ARDL Bounds Test | Johansen Method |
|---------|------------------|-----------------|
| **Order of integration** | I(0), I(1), or mixed | All must be I(1) |
| **Pre-testing required** | No (only check not I(2)) | Yes (must confirm I(1)) |
| **Estimation** | Single equation | System (VAR) |
| **Complexity** | Simpler | More complex |
| **Small samples** | Works well (n≥30) | Needs larger samples |
| **Different lags** | Yes, per variable | No, same for all |
| **Endogeneity** | Can handle with lags | Better for simultaneous equations |

## When to Use ARDL {.smaller}

**ARDL is preferred when:**

::: {.incremental}
1. **Mixed integration orders** (some I(0), some I(1))
2. **Small sample size** (< 100 observations)
3. **Single equation focus** (one dependent variable of interest)
4. **Different dynamics** across variables
5. **Simpler interpretation** is desired
6. **Preliminary analysis** before full system estimation
:::

## When to Use Johansen {.smaller}

**Johansen is preferred when:**

::: {.incremental}
1. **Multiple cointegrating relationships** suspected
2. **All variables are I(1)** (confirmed)
3. **System-wide analysis** needed
4. **Simultaneous determination** (multiple endogenous variables)
5. **Large sample** available
6. **Bidirectional causality** important
:::

::: {.callout-note}
Many researchers use **both methods** as robustness checks!
:::

# Practical Example

## Real Data Example {.smaller}

Let's analyze the relationship between GDP, Investment, and Government Spending using Pakistani data:

```{r}
#| eval: false

# Load data (example - use your own data)
library(readr)
pak_data <- read_csv("pakistan_macro.csv")

# Variables:
# - gdp: Real GDP (in billions)
# - investment: Gross Fixed Capital Formation
# - govt_spending: Government Expenditure

# Take logs for elasticity interpretation
pak_data <- pak_data %>%
  mutate(
    log_gdp = log(gdp),
    log_inv = log(investment),
    log_govt = log(govt_spending)
  )
```

## Unit Root Tests {.smaller}

```{r}
#| eval: false

library(urca)

# ADF tests on levels
adf_gdp <- ur.df(pak_data$log_gdp, type = "trend", lags = 4)
adf_inv <- ur.df(pak_data$log_inv, type = "trend", lags = 4)
adf_govt <- ur.df(pak_data$log_govt, type = "trend", lags = 4)

# Summary
summary(adf_gdp)
summary(adf_inv)
summary(adf_govt)

# Check first differences if needed
adf_d_gdp <- ur.df(diff(pak_data$log_gdp), type = "trend", lags = 4)
summary(adf_d_gdp)
```

**Expected result:** Variables are I(1), good for ARDL!

## Estimate ARDL Model {.smaller}

```{r}
#| eval: false

# Auto-select optimal ARDL model
ardl_pak <- auto_ardl(
  log_gdp ~ log_inv + log_govt,
  data = pak_data,
  max_order = 4,
  selection = "BIC"  # Use BIC for parsimony
)

# View selected order
ardl_pak$order
# Example output: (2, 1, 1) = ARDL(2,1,1)

# Full model summary
summary(ardl_pak)
```

## Bounds Test Results {.smaller}

```{r}
#| eval: false

# Perform bounds test
bounds_result <- bounds_f_test(ardl_pak, case = 3)

# View results
print(bounds_result)

# Example output:
#           F-statistic  Lower.bound  Upper.bound
# 10%           5.234        2.72         3.77
# 5%            5.234        3.23         4.35
# 1%            5.234        4.29         5.61
```

**Interpretation:** F = 5.234 > Upper bound at 5% (4.35)  
→ **Strong evidence of cointegration!**

## Long-Run Results {.smaller}

```{r}
#| eval: false

# Extract long-run coefficients
lr_pak <- multipliers(ardl_pak, type = "lr")

# View results
summary(lr_pak)

# Example output:
#              Estimate  Std.Error  t-value  Pr(>|t|)
# log_inv        0.456      0.089    5.123    0.000 ***
# log_govt       0.234      0.067    3.493    0.001 **
```

**Economic Interpretation:**

- 1% increase in investment → 0.456% increase in GDP (long-run)
- 1% increase in govt spending → 0.234% increase in GDP (long-run)

## ECM Results {.smaller}

```{r}
#| eval: false

# Estimate ECM
ecm_pak <- ecm(ardl_pak)

# View results
summary(ecm_pak)

# Example output:
#              Estimate  Std.Error  t-value  Pr(>|t|)
# ect           -0.428      0.098   -4.367    0.000 ***
# d(log_inv)     0.312      0.076    4.105    0.000 ***
# d(log_govt)    0.189      0.054    3.500    0.001 **
```

**Interpretation:**

- ECT = -0.428: About 43% of disequilibrium corrected per period
- Speed of adjustment: $1/0.428 \approx 2.3$ periods to equilibrium
- Short-run effects also significant

## Diagnostic Checks {.smaller}

```{r}
#| eval: false

# Load required package
library(lmtest)

# 1. Breusch-Godfrey Test (Serial Correlation)
bgtest(ardl_pak, order = 4)
# H0: No serial correlation

# 2. Breusch-Pagan Test (Heteroskedasticity)
bptest(ardl_pak)
# H0: Homoskedastic errors

# 3. Jarque-Bera Test (Normality)
jarque.bera.test(residuals(ardl_pak))
# H0: Errors are normally distributed

# 4. CUSUM Test (Stability)
cusum_test <- stability(ardl_pak, type = "cusum")
plot(cusum_test)
```

## Visualizing the Relationship {.smaller}

```{r}
#| eval: false

# Plot actual vs fitted
library(ggplot2)

fitted_vals <- fitted(ardl_pak)
actual_vals <- pak_data$log_gdp[1:length(fitted_vals)]

plot_df <- data.frame(
  time = 1:length(actual_vals),
  actual = actual_vals,
  fitted = fitted_vals
)

ggplot(plot_df, aes(x = time)) +
  geom_line(aes(y = actual, color = "Actual"), size = 1) +
  geom_line(aes(y = fitted, color = "Fitted"), size = 1, linetype = "dashed") +
  labs(title = "ARDL Model: Actual vs Fitted Values",
       subtitle = "Log GDP in Pakistan",
       x = "Time Period", y = "Log GDP",
       color = "Series") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Advanced Topics

## ARDL with Structural Breaks {.smaller}

If you suspect structural breaks (e.g., policy changes, crises):

```{r}
#| eval: false

# Add dummy variables
pak_data$crisis_dummy <- ifelse(pak_data$year >= 2008 & 
                                pak_data$year <= 2010, 1, 0)

# Estimate ARDL with dummy
ardl_break <- auto_ardl(
  log_gdp ~ log_inv + log_govt + crisis_dummy,
  data = pak_data,
  max_order = 4
)

# Check if dummy is significant
summary(ardl_break)
```

## ARDL with Non-linear Effects {.smaller}

Include squared terms for non-linear relationships:

```{r}
#| eval: false

# Create squared term
pak_data$log_inv_sq <- pak_data$log_inv^2

# Estimate model
ardl_nonlin <- auto_ardl(
  log_gdp ~ log_inv + log_inv_sq + log_govt,
  data = pak_data,
  max_order = 4
)

summary(ardl_nonlin)
```

**Test for non-linearity:** If $\beta_{inv^2}$ is significant, relationship is non-linear

## NARDL: Non-linear ARDL {.smaller}

Account for **asymmetric effects** (positive vs negative changes):

```{r}
#| eval: false

library(nardl)

# Separate positive and negative changes
nardl_model <- nardl(
  log_gdp ~ log_inv + log_govt,
  data = pak_data,
  p = 2, q = 1  # ARDL(2,1) structure
)

# Test for asymmetry
asymmetry_test <- nardl_model$pss

# View results
summary(nardl_model)
```

::: {.callout-note}
NARDL is useful when effects differ in expansions vs recessions
:::

# Common Mistakes & Tips

## Common Mistakes {.smaller}

::: {.incremental}
1. **Not checking for I(2) variables**
   - Always test! I(2) invalidates bounds test
   
2. **Ignoring diagnostic tests**
   - Serial correlation biases standard errors
   - Check residuals carefully
   
3. **Over-fitting with too many lags**
   - Use information criteria (AIC/BIC)
   - Keep model parsimonious
   
4. **Misinterpreting ECT sign**
   - Must be negative for valid cointegration
   - Positive ECT = explosive process
:::

## Common Mistakes (continued) {.smaller}

::: {.incremental}
5. **Not testing coefficient restrictions**
   - Test long-run coefficients significance
   - Use Wald tests when needed

6. **Ignoring structural breaks**
   - Check data for known shocks/policy changes
   - Use CUSUM/CUSUMSQ tests

7. **Wrong case selection in bounds test**
   - Case 1: No intercept, no trend
   - Case 2: Restricted intercept, no trend
   - Case 3: Unrestricted intercept, no trend ← Most common
   - Case 4: Unrestricted intercept, restricted trend
   - Case 5: Unrestricted intercept, unrestricted trend
:::

## Best Practices {.smaller}

::: {.incremental}
1. **Always plot your data first**
   - Check for trends, breaks, outliers
   
2. **Use log transformations when appropriate**
   - For multiplicative relationships
   - To interpret coefficients as elasticities
   
3. **Report multiple model specifications**
   - Different lag lengths
   - With/without additional controls
   
4. **Conduct robustness checks**
   - Compare ARDL with Johansen
   - Try different lag selection criteria
   
5. **Interpret economically, not just statistically**
   - Do magnitudes make sense?
   - Are signs consistent with theory?
:::

## Reporting ARDL Results {.smaller}

**In your paper/thesis, report:**

::: {.incremental}
1. **Unit root test results** (confirm not I(2))
2. **Selected ARDL specification** and lag selection criteria
3. **Bounds test results** with F-statistic and decision
4. **Long-run coefficients** with standard errors
5. **Short-run coefficients** and ECT from ECM
6. **Diagnostic test results** (serial correlation, heteroskedasticity, normality, stability)
7. **Economic interpretation** of all key coefficients
:::

::: {.callout-tip}
Create a comprehensive table with all results for easy reference!
:::

# Conclusion & Resources

## Summary {.smaller}

**Key Takeaways:**

::: {.incremental}
1. ARDL is **flexible** - works with I(0), I(1), or mixed
2. **Bounds testing** avoids unit root pre-testing debates
3. **Simpler** than Johansen for single equation models
4. Provides both **short-run dynamics** and **long-run equilibrium**
5. **R implementation** is straightforward with ARDL package
6. Always check **diagnostics** and **stability**
7. **Economic interpretation** matters most!
:::

## When to Use What? {.scrollable}

**Quick Decision Guide:**

::: {.columns}
::: {.column width="60%"}

```{mermaid}
%%| fig-width: 8
flowchart TD
    A[Start: Check Your Data] --> B{Variables <br/> mixed I0/I1?}
    B -->|Yes| C[Use ARDL]
    B -->|No, All I1| D{Small sample <br/> n < 80?}
    D -->|Yes| C
    D -->|No| E{Single equation <br/> focus?}
    E -->|Yes| F[ARDL preferred <br/> Simpler]
    E -->|No| G{Multiple <br/> cointegrating <br/> relations?}
    G -->|Yes| H[Use Johansen]
    G -->|No| F
    
    style C fill:#90EE90
    style H fill:#FFB6C1
    style F fill:#90EE90
    
    C --> I[ARDL Bounds Test]
    H --> J[Johansen Test]
    F --> K[Either works <br/> Use both for robustness]
```

:::

::: {.column width="40%"}

**Key Decision Points:**

1. **Mixed Integration?**  
   - I(0) and I(1) mixed → ARDL only

2. **Sample Size?**  
   - Small (n<80) → ARDL better
   - Large (n>100) → Both work

3. **Research Focus?**  
   - Single equation → ARDL simpler
   - System dynamics → Johansen

4. **Multiple Cointegration?**  
   - Multiple relations → Johansen
   - Single relation → ARDL

:::
:::

::: {.callout-tip}
**Best Practice:** Use both methods as robustness checks when possible!
:::

## R Packages Summary {.smaller}

**Essential packages:**

```{r}
#| eval: false

# ARDL analysis
install.packages("ARDL")      # Main package
install.packages("dynamac")   # Alternative

# Unit root tests
install.packages("urca")      # Comprehensive tests
install.packages("tseries")   # ADF, PP, KPSS

# Diagnostics
install.packages("lmtest")    # LM tests
install.packages("car")       # VIF, other tests

# Visualization
install.packages("ggplot2")   # Publication-quality plots
```

## Additional Resources {.smaller}

**Key Papers:**

1. Pesaran, M. H., Shin, Y., & Smith, R. J. (2001). Bounds testing approaches to the analysis of level relationships. *Journal of Applied Econometrics*, 16(3), 289-326.

2. Pesaran, M. H., & Shin, Y. (1999). An autoregressive distributed lag modelling approach to cointegration analysis.

**Online Resources:**

- ARDL R package documentation: `help(package = "ARDL")`
- Econometrics tutorials: [https://www.econometrics-with-r.org/](https://www.econometrics-with-r.org/)
- Time series in R: [https://otexts.com/fpp3/](https://otexts.com/fpp3/)

## Practice Exercise {.smaller}

**Try this on your own:**

1. Load any macroeconomic dataset with 3-4 variables
2. Test for unit roots (ensure not I(2))
3. Estimate an ARDL model using `auto_ardl()`
4. Conduct bounds test for cointegration
5. If cointegrated:
   - Estimate long-run relationship
   - Estimate ECM
   - Interpret the speed of adjustment
6. Run diagnostic tests
7. Visualize actual vs fitted values

::: {.callout-tip}
Start with Pakistani data from World Bank or State Bank of Pakistan!
:::

## Questions? {.center}

::: {.r-fit-text}
Thank you!
:::

**Contact:**

- Email: [your.email@university.edu](mailto:your.email@university.edu)
- Office Hours: [Your hours]

::: {.callout-note}
## Remember:
Practice makes perfect. Start with simple examples and gradually increase complexity!
:::

---

### References {.smaller}

1. Pesaran, M. H., Shin, Y., & Smith, R. J. (2001). Bounds testing approaches to the analysis of level relationships. *Journal of Applied Econometrics*, 16(3), 289-326.

2. Pesaran, M. H., & Shin, Y. (1999). An autoregressive distributed lag modelling approach to cointegration analysis. *Econometrics and Economic Theory in the 20th Century: The Ragnar Frisch Centennial Symposium*, 371-413.

3. Nkoro, E., & Uko, A. K. (2016). Autoregressive Distributed Lag (ARDL) cointegration technique: application and interpretation. *Journal of Statistical and Econometric Methods*, 5(4), 63-91.

4. Jordan, S., & Philips, A. Q. (2018). Cointegration testing and dynamic simulations of autoregressive distributed lag models. *The Stata Journal*, 18(4), 902-923.
