---
title: "Remittances in Pakistan: Trends and Forecasting"
subtitle: "Live data from the SBP EasyData API"
author: "Zahid Asghar"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: show
execute:
  warning: false
  message: false
editor_options:
  chunk_output_type: console
---

## Introduction

This post looks at workers' remittances flowing into Pakistan, broken down by the
country the money is sent from. Rather than working off a downloaded spreadsheet
that goes stale the moment it is saved, the data here comes straight from the
State Bank of Pakistan's EasyData API, so re-rendering the document refreshes the
numbers.

Two things are worth knowing about the series before reading too much into the
country split. SBP records remittances by the *original country of the remitter*
only from July 2019 onwards, so the earlier country breakdown is compiled on a
different basis. And from July 2022 the series also absorbs encashment and profit
on FEBCs and FCBCs. Values are in million US dollars.

## Setup

```{r setup}
library(tidyverse)   # dplyr, ggplot2, tidyr, purrr, readr, stringr
library(httr2)       # API requests
library(janitor)     # clean_names()
library(fpp3)        # loads tsibble, feasts and fable together
library(plotly)      # interactive plots

theme_set(theme_minimal(base_size = 12))
```

Note that `fpp3` already brings in `tsibble`, `feasts` and `fable`. The older
`forecast` and `xts` packages are deliberately left out: loading `forecast`
alongside `fable` masks `forecast()` and `accuracy()`, which is the source of a
lot of confusing errors. Everything below is done the fable way.

## Talking to EasyData

The API follows one pattern:

```
https://easydata.sbp.org.pk/api/v1/series/{SERIES_ID}/data?api_key={KEY}&start_date=&end_date=&format=csv
```

The key lives in `~/.Renviron` as `SBP_EASYDATA_KEY` and never appears in this
document. If the next chunk complains that the key is missing, run
`usethis::edit_r_environ()`, add the line, make sure the file ends with a blank
line, and restart R.

```{r api-helpers}
sbp_raw <- function(series_id,
                    start_date = "2010-07-01",
                    end_date   = Sys.Date(),
                    key        = Sys.getenv("SBP_EASYDATA_KEY")) {

  if (!nzchar(key)) {
    stop("SBP_EASYDATA_KEY not found. Run usethis::edit_r_environ(), ",
         "add the key, end the file with a blank line, then restart R.")
  }

  request("https://easydata.sbp.org.pk/api/v1/series") |>
    req_url_path_append(series_id, "data") |>
    req_url_query(
      api_key    = key,
      start_date = as.character(start_date),
      end_date   = as.character(end_date),
      format     = "csv"
    ) |>
    req_timeout(60) |>
    req_retry(max_tries = 3) |>
    req_perform() |>
    resp_body_string() |>
    I() |>
    read_csv(show_col_types = FALSE) |>
    clean_names()
}

# The CSV carries several metadata columns; pick them out by pattern rather
# than by position, so a change in column order does not break the pipeline.
pick_col <- function(df, pattern) {
  hit <- names(df)[str_detect(names(df), pattern)]
  if (length(hit) == 0) NA_character_ else hit[1]
}

sbp_series <- function(series_id, ...) {
  raw <- sbp_raw(series_id, ...)

  date_col  <- pick_col(raw, "observation_date|^date$|period")
  value_col <- pick_col(raw, "observation_value|^value$")

  if (is.na(date_col) || is.na(value_col)) {
    stop("Unexpected columns for ", series_id, ": ",
         paste(names(raw), collapse = ", "))
  }

  raw |>
    transmute(
      date         = as.Date(.data[[date_col]]),
      remittances  = as.numeric(.data[[value_col]])
    ) |>
    filter(!is.na(date)) |>
    arrange(date)
}
```

## Finding the country series codes

Workers' remittances sit in the monthly dataset `TS_GP_BOP_WR_M`, with one series
per source country keyed as `TS_GP_BOP_WR_M.WRnnnn`. Rather than typing codes out
of the portal by hand, this chunk discovers them: it first tries the dataset-level
endpoint, and if that is not available it probes the series codes and reads each
country's name out of the returned metadata.

Run it once. It writes a lookup table to disk, and everything afterwards reads
from that.

```{r discover-series}
#| eval: false

lookup_file <- "data/sbp_wr_series.csv"
dir.create("data", showWarnings = FALSE, recursive = TRUE)

# Attempt 1: whole dataset in one call, if SBP exposes it on your key
dataset_try <- try(
  request("https://easydata.sbp.org.pk/api/v1/dataset") |>
    req_url_path_append("TS_GP_BOP_WR_M", "data") |>
    req_url_query(api_key = Sys.getenv("SBP_EASYDATA_KEY"), format = "csv") |>
    req_timeout(120) |>
    req_perform() |>
    resp_body_string() |>
    I() |>
    read_csv(show_col_types = FALSE) |>
    clean_names(),
  silent = TRUE
)

if (!inherits(dataset_try, "try-error")) {

  print(names(dataset_try))   # inspect, then keep the code + name columns
  glimpse(dataset_try)

} else {

  # Attempt 2: probe the series codes one at a time.
  probe <- function(code) {
    id  <- paste0("TS_GP_BOP_WR_M.", code)
    out <- try(sbp_raw(id, start_date = "2024-01-01", end_date = "2024-06-30"),
               silent = TRUE)
    Sys.sleep(0.3)   # be polite to the server
    if (inherits(out, "try-error") || nrow(out) == 0) return(NULL)

    name_col <- pick_col(out, "series_name|series_title|^name$|title|description")
    tibble(
      code        = code,
      series_id   = id,
      series_name = if (is.na(name_col)) NA_character_ else as.character(out[[name_col]][1])
    )
  }

  codes <- sprintf("WR%04d", seq(10, 600, by = 10))
  series_lookup <- map_dfr(codes, probe)

  write_csv(series_lookup, lookup_file)
  print(series_lookup, n = Inf)
}
# Single known series, no error suppression
test <- sbp_raw("TS_GP_BOP_WR_M.WR0320",
                start_date = "2024-01-01",
                end_date   = "2024-06-30")

names(test)
glimpse(test)
```

Open `data/sbp_wr_series.csv`, read the country names, and keep the ones you want.
The next chunk builds `series_map` from that file, so no codes are hard-coded here.

```{r build-series-map}
lookup_file <- "data/sbp_wr_series.csv"

stopifnot(
  "Run the discovery chunk first to create data/sbp_wr_series.csv" =
    file.exists(lookup_file)
)

series_lookup <- read_csv(lookup_file, show_col_types = FALSE) |>
  filter(!is.na(series_name)) |>
  mutate(
    # Tidy the SBP label into a short country name for plotting
    country = series_name |>
      str_remove(regex("^workers'?\\s*remittances\\s*[-:,]?\\s*", ignore_case = TRUE)) |>
      str_remove(regex("\\s*\\(.*\\)$")) |>
      str_squish()
  )

series_map <- set_names(series_lookup$series_id, series_lookup$country)

length(series_map)
head(names(series_map), 20)
```

## Pulling and caching the data

Every render would otherwise hit the API once per country. The cache keeps a local
copy for a week, and falls back to it if SBP is unreachable mid-render.

```{r fetch-data}
cache_file <- "data/remittances_sbp.rds"
dir.create(dirname(cache_file), showWarnings = FALSE, recursive = TRUE)

fetch_all <- function() {
  imap_dfr(series_map, \(id, country) {
    Sys.sleep(0.2)
    sbp_series(id) |> mutate(country = country)
  })
}

cache_fresh <- file.exists(cache_file) &&
  difftime(Sys.time(), file.mtime(cache_file), units = "days") < 7

rmt_long <- if (cache_fresh) {
  read_rds(cache_file)
} else {
  out <- try(fetch_all(), silent = TRUE)
  if (inherits(out, "try-error")) {
    if (!file.exists(cache_file)) stop(out)
    message("SBP API unreachable, using cached copy from ", file.mtime(cache_file))
    read_rds(cache_file)
  } else {
    write_rds(out, cache_file)
    out
  }
}

rmt_long <- rmt_long |>
  filter(!is.na(remittances)) |>
  mutate(month = yearmonth(date))

glimpse(rmt_long)
range(rmt_long$date)
```

To force a refresh before publishing: `file.remove("data/remittances_sbp.rds")`.

### As a tsibble

```{r make-tsibble}
ts_data <- rmt_long |>
  as_tsibble(key = country, index = month) |>
  fill_gaps()          # implicit gaps break the seasonal models

has_gaps(ts_data) |> filter(.gaps) |> head()
```

`fill_gaps()` matters here: countries enter the series at different dates, and
fable refuses to fit seasonal models on a tsibble with implicit gaps.

### A wide copy, if you want one

```{r wide-copy}
rmt_wide <- rmt_long |>
  select(date, country, remittances) |>
  pivot_wider(names_from = country, values_from = remittances) |>
  arrange(date)

dim(rmt_wide)
```

## Trends by country

```{r plot-trends}
#| fig-height: 6

total_label <- names(series_map)[str_detect(names(series_map),
                                            regex("total|all countries", ignore_case = TRUE))]

top_countries <- rmt_long |>
  filter(!country %in% total_label) |>
  group_by(country) |>
  summarise(mean_flow = mean(remittances, na.rm = TRUE), .groups = "drop") |>
  slice_max(mean_flow, n = 6) |>
  pull(country)

p_trend <- rmt_long |>
  filter(country %in% top_countries) |>
  ggplot(aes(date, remittances, colour = country)) +
  geom_line(linewidth = 0.6) +
  labs(
    title = "Workers' remittances to Pakistan by source country",
    x = NULL, y = "US$ million", colour = NULL,
    caption = "Source: State Bank of Pakistan, EasyData"
  )

ggplotly(p_trend)
```

## Country shares

The original version filtered on `share > 0.08`, which drops individual *months*
rather than countries and leaves broken lines. Ranking countries by their average
share and then plotting the full series for the leaders is what was intended.

```{r shares}
#| fig-height: 5

totals <- if (length(total_label) > 0) {
  rmt_long |> filter(country %in% total_label) |> select(date, total = remittances)
} else {
  rmt_long |> group_by(date) |> summarise(total = sum(remittances, na.rm = TRUE))
}

rmt_share <- rmt_long |>
  filter(!country %in% total_label) |>
  left_join(totals, by = "date") |>
  mutate(share = remittances / total) |>
  filter(is.finite(share))

share_leaders <- rmt_share |>
  group_by(country) |>
  summarise(mean_share = mean(share, na.rm = TRUE), .groups = "drop") |>
  filter(mean_share > 0.05) |>
  pull(country)

p_share <- rmt_share |>
  filter(country %in% share_leaders) |>
  ggplot(aes(date, share, colour = country)) +
  geom_line(linewidth = 0.6) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Share of total remittances by source country",
    x = NULL, y = "Share of total", colour = NULL
  )

ggplotly(p_share)
```

## Seasonality and decomposition

```{r decomposition}
#| fig-height: 6

focus <- top_countries[1]

ts_data |>
  filter(country == focus) |>
  model(STL(remittances ~ season(window = "periodic"))) |>
  components() |>
  autoplot() +
  labs(title = paste("STL decomposition:", focus))
```

```{r seasonal-subseries}
#| fig-height: 5

ts_data |>
  filter(country == focus, !is.na(remittances)) |>
  gg_subseries(remittances) +
  labs(title = paste("Seasonal pattern:", focus), y = "US$ million")
```

The Ramadan and Eid effect is usually visible here, and it moves through the
calendar year, which is one reason a purely seasonal-naive model struggles.

## Forecasting

Three models fitted side by side, on the countries that matter most, with a
holdout period so the comparison means something.

```{r fit-models}
model_set <- ts_data |>
  filter(country %in% top_countries)

train <- model_set |> filter(month < max(month) - 23)

fits <- train |>
  model(
    snaive = SNAIVE(remittances ~ lag("year")),
    ets    = ETS(remittances),
    arima  = ARIMA(remittances)
  )

fits
```

```{r accuracy}
fc_test <- fits |> forecast(h = "24 months")

fc_test |>
  accuracy(model_set) |>
  select(country, .model, RMSE, MAE, MAPE, MASE) |>
  arrange(country, RMSE) |>
  print(n = Inf)
```

```{r forecast-plot}
#| fig-height: 7

final_fits <- model_set |>
  model(
    snaive = SNAIVE(remittances ~ lag("year")),
    ets    = ETS(remittances),
    arima  = ARIMA(remittances)
  )

fc <- final_fits |> forecast(h = "12 months")

fc |>
  autoplot(model_set |> filter(date >= max(date) - years(5)), level = 80) +
  facet_wrap(~ country, scales = "free_y", ncol = 2) +
  labs(
    title = "12-month forecasts by model",
    x = NULL, y = "US$ million"
  )
```

### Total remittances

```{r forecast-total}
#| fig-height: 5

total_series <- if (length(total_label) > 0) {
  ts_data |> filter(country %in% total_label)
} else {
  rmt_long |>
    filter(!country %in% total_label) |>
    group_by(month) |>
    summarise(remittances = sum(remittances, na.rm = TRUE), .groups = "drop") |>
    as_tsibble(index = month)
}

total_series |>
  model(
    ets   = ETS(remittances),
    arima = ARIMA(remittances)
  ) |>
  forecast(h = "12 months") |>
  autoplot(total_series |> filter(month >= max(month) - 60), level = 80) +
  labs(title = "Total remittances: 12-month forecast", x = NULL, y = "US$ million")
```

### Residual checks

```{r residuals}
#| fig-height: 5

final_fits |>
  filter(country == focus) |>
  select(arima) |>
  gg_tsresiduals() +
  labs(title = paste("ARIMA residuals:", focus))
```

## Conclusion

The data now arrives live from SBP rather than from a file, the country labels come
from SBP's own metadata rather than from column names, and the forecasting is done
entirely within the fable framework so models can be compared on a common holdout.

Natural extensions: aggregate the country series with `reconcile()` so the parts
sum to the published total, add oil prices or the PKR/USD rate as regressors in a
dynamic regression, and test whether the July 2019 change in compilation basis
shows up as a structural break in the country shares.
