---
title: "Remittances in Pakistan: Trends and Forecasting"
author: "Zahid Asghar"
date: "2024-12-XX"
format: 
  html:
    toc: true
    toc-depth: 2
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

## Introduction

This blog post explores remittance data for Pakistan—from various countries—to understand historical trends and forecast future remittance flows. We will:
- Load and clean the remittance data
- Visualize remittance trends by country
- Calculate each country’s share of total remittances
- Forecast remittance series using several time series models

We will use R packages such as **fable**, **fpp3**, **tsibble**, **dplyr**, **ggplot2**, **forecast**, and **plotly** for our analysis.

---

## Load Libraries and Data

```{r}
#| eval: false
# Load necessary libraries
library(fable)       # Forecasting tools
library(fpp3)        # Time series analysis and forecasting
library(tsibble)     # Time series data manipulation
library(dplyr)       # Data manipulation
library(ggplot2)     # Data visualization
library(xts)         # Time series manipulation
library(lubridate)   # Date manipulation
library(tidyr)       # Data transformation
library(forecast)    # Forecasting functions
library(plotly)      # Interactive plots

# Load remittance data (up to December 2024 from Karandaz)
rmt <- read_csv("data/remittances_122024.csv")

# Inspect data
rmt |> glimpse()

# Remove the first row if it is not needed and convert columns (except Period) to numeric
rmt <- rmt[-1,] |> 
  mutate(across(-Period, as.numeric))

# Create a date column: generate dates from December 2024 to July 2010 (monthly)
dates <- seq(as.Date("2024-12-01"), as.Date("2010-07-01"), by = "-1 month")
rmt$date <- dates

# Reorder columns so that date is first
rmt <- rmt %>% select(-Period) %>% select(date, everything())
```


```{r sbp-remittances}
library(tidyverse)
library(httr2)
library(janitor)
library(tsibble)

sbp_series <- function(series_id,
                       start_date = "2010-07-01",
                       end_date   = Sys.Date(),
                       key        = Sys.getenv("SBP_EASYDATA_KEY")) {

  if (!nzchar(key)) {
    stop("SBP_EASYDATA_KEY not found. Run usethis::edit_r_environ(), add the key, restart R.")
  }

  request("https://easydata.sbp.org.pk/api/v1/series") |>
    req_url_path_append(series_id, "data") |>
    req_url_query(
      api_key    = key,
      start_date = as.character(start_date),
      end_date   = as.character(end_date),
      format     = "csv"
    ) |>
    req_timeout(60) |>
    req_retry(max_tries = 3) |>
    req_perform() |>
    resp_body_string() |>
    I() |>
    read_csv(show_col_types = FALSE) |>
    clean_names() |>
    transmute(
      date  = as.Date(observation_date),
      value = as.numeric(observation_value)
    ) |>
    arrange(date)
}

# Replace the WRxxxx codes with the keys you pulled from the EasyData portal
series_map <- c(
  "Saudi Arabia" = "TS_GP_BOP_WR_M.WRxxxx",
  "UAE"          = "TS_GP_BOP_WR_M.WRxxxx",
  "UK"           = "TS_GP_BOP_WR_M.WRxxxx",
  "USA"          = "TS_GP_BOP_WR_M.WRxxxx",
  "Total"        = "TS_GP_BOP_WR_M.WRxxxx"
)

cache_file <- "data/remittances_sbp.rds"
dir.create(dirname(cache_file), showWarnings = FALSE, recursive = TRUE)

fetch_all <- function() {
  imap_dfr(series_map, \(id, country) sbp_series(id) |> mutate(country = country))
}

cache_fresh <- file.exists(cache_file) &&
  difftime(Sys.time(), file.mtime(cache_file), units = "days") < 7

rmt_long <- if (cache_fresh) {
  read_rds(cache_file)
} else {
  out <- try(fetch_all(), silent = TRUE)
  if (inherits(out, "try-error")) {
    if (!file.exists(cache_file)) stop(out)
    message("SBP API unreachable, falling back to cached copy from ", file.mtime(cache_file))
    read_rds(cache_file)
  } else {
    write_rds(out, cache_file)
    out
  }
}

rmt_ts <- rmt_long |>
  mutate(month = yearmonth(date)) |>
  as_tsibble(index = month, key = country)

# Wide version, if you prefer the old layout
rmt <- rmt_long |>
  pivot_wider(names_from = country, values_from = value) |>
  arrange(date)
```



---

## Data Transformation and Visualization 

After loading and cleaning the data, we will transform it into a long format for easier plotting by country. We will then visualize remittance trends by country and calculate each country's share of total remittances.

### Wide to Long Format

```{r}
# Convert data from wide to long format for easier plotting by country
rmt_long <- rmt %>% 
  pivot_longer(-date, names_to = "country", values_to = "remittances")

# List distinct countries (excluding "Total" if needed)
rmt_long |> distinct(country)
```

### Remittances by Country

```{r}
# Plot remittances for each country (excluding the "Total" column)
rmt_long |> 
  filter(country != "Total") |> 
  ggplot(aes(x = date, y = remittances, color = country)) +
  geom_line() +
  labs(title = "Remittances by Country", x = "Date", y = "Remittances") +
  theme_minimal()
```

### Calculate Country Share

```{r}
# Calculate each country's share of total remittances
rmt_share <- rmt |> 
  mutate(across(-date, ~ . / Total)) 

# Convert to long format for visualization
rmt_share_long <- rmt_share %>% 
  pivot_longer(-date, names_to = "country", values_to = "share")

# Plot share for countries with a share greater than 8%
rmt_share_long |> 
  filter(country != "Total", share > 0.08) |>
  ggplot(aes(x = date, y = share, color = country)) +
  geom_line() +
  labs(title = "Country Share of Total Remittances", x = "Date", y = "Share") +
  theme_minimal() -> p1
ggplotly(p1)
```

### Interactive Plot 



```{r}
# Create an interactive plot using plotly (excluding "Total")
p <- rmt_long |> 
  filter(country != "Total") |>
  ggplot(aes(x = date, y = remittances, color = country)) +
  geom_line() +
  labs(title = "Remittances by Country", x = "Date", y = "Remittances") +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p)
```

---

## Time Series Forecasting

### Preparing the Data as a tsibble 

In this section, we will convert the long-format data into a tsibble object, which is a time series data structure in R. We will then use the Seasonal Naive (SNAIVE) model to forecast total remittances and selected countries.

```{r}
# Create a year-month index for tsibble conversion
rmt_long <- rmt_long |> 
  mutate(month_year = yearmonth(date))

# Convert to a tsibble, keyed by country
ts_data <- as_tsibble(rmt_long, key = country, index = month_year)
```

### Forecasting with SNAIVE

We will use the Seasonal Naive (SNAIVE) model to forecast total remittances and selected countries. The SNAIVE model uses the last observed value from the same season as the forecast.

```{r}
# Forecast Total remittances using a seasonal naive (SNAIVE) model
ts_data |> 
  filter(country == "Total") |> 
  model(SNAIVE(remittances ~ lag(12))) |> 
  forecast(h = "12 months") |> 
  autoplot(ts_data) +
  labs(title = "12-Month Forecast for Total Remittances")
```

Forecasts for selected countries (e.g., Saudi Arabia, United Arab Emirates, USA, United Kingdom, EU Countries):

```{r}
countries <- c("Saudi.Arabia", "United.Arab.Emirates", "USA", "United.Kingdom", "EU.Countries")
for (ctry in countries) {
  ts_data |> 
    filter(country == ctry) |> 
    model(SNAIVE(remittances ~ lag(12))) |> 
    forecast(h = "12 months") |> 
    autoplot(ts_data) +
    labs(title = paste("12-Month Forecast for", ctry)) -> p
  print(p)
}
```

### Decomposition and Alternative Models

```{r}
# Decomposition for one selected country (e.g., Saudi Arabia)
saudi_ts <- ts_data |> 
  filter(country == "Saudi.Arabia") |> 
  select(month_year, remittances) |> 
  as_tsibble(index = month_year)

decomp <- saudi_ts |> model(STL(remittances ~ season(window = "periodic"))) |> components()
autoplot(decomp) + ggtitle("STL Decomposition for Saudi Arabia Remittances") + theme_minimal()

# Convert the Saudi Arabia series to a ts object for exponential smoothing methods
saudi_rem_ts <- ts(saudi_ts$remittances, 
                   start = c(year(min(saudi_ts$month_year)), month(min(saudi_ts$month_year))),
                   frequency = 12)

# Forecast using Simple Exponential Smoothing, Holt's and Holt-Winters methods
autoplot(forecast(ses(saudi_rem_ts, h = 12))) + ggtitle("SES Forecast")
autoplot(forecast(holt(saudi_rem_ts, h = 12))) + ggtitle("Holt's Linear Trend Forecast")
autoplot(forecast(hw(saudi_rem_ts, h = 12))) + ggtitle("Holt-Winters Forecast")
```

### ARIMA Model

```{r}
# Fit an ARIMA model for Saudi Arabia remittances
fit_ARIMA <- auto.arima(saudi_rem_ts, stepwise = FALSE, approximation = FALSE)
summary(fit_ARIMA)

# Forecast using the ARIMA model
autoplot(forecast(fit_ARIMA, h = 24)) + ggtitle("ARIMA Forecast for Saudi Arabia Remittances")
```

---

## Conclusion

In this post, we:
- Loaded and cleaned remittance data for Pakistan
- Transformed the data to visualize trends and country shares
- Explored forecasting techniques including seasonal naive models, exponential smoothing, and ARIMA

This analysis helps illustrate the dynamics of remittance flows and provides insights that may inform policy and economic forecasting. Future work might involve further model comparisons and the inclusion of exogenous variables.

Happy forecasting!

