---
title: "Remittances Forecasting Analysis for `r params$country`"
subtitle: "6-Month Ahead Forecast with Multiple Methods"
format: 
  html:
    embed_resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
  pdf: 
    toc: true
  docx: default
execute:
  messages: false
  warning: false
  echo: false
params:
  country: "Total"
  forecast_horizon: 6
editor_options: 
  chunk_output_type: console
---
```{r setup}
#| include: false
library(fable)
library(fpp3)
library(tsibble)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(gt)
library(patchwork)
library(kableExtra)

# Set ggplot theme
theme_set(theme_minimal(base_size = 12))
```

# Executive Summary

This report provides a comprehensive 6-month forecast of remittances for **`r params$country`** using multiple time series forecasting methods. We employ:

- **ARIMA** (AutoRegressive Integrated Moving Average)
- **ETS** (Exponential Smoothing State Space)
- **SNAIVE** (Seasonal Naive with 12-month lag)
- **Ensemble** (Combined forecast)

The best-performing model is selected based on accuracy metrics and used for final predictions.

---

# 1. Data Loading and Preparation
```{r data-load}
# Load data
rmt <- read.csv("data/remittances_sep25.csv")

# Clean data
rmt <- rmt[-1,]
rmt[, -1] <- lapply(rmt[, -1], as.numeric)

# Create date sequence
dates <- seq(as.Date("2025-09-01"), as.Date("2010-07-01"), by = "-1 month")
rmt$date <- dates

# Reorder columns
rmt <- rmt %>% select(-Period) %>% select(date, everything())

# Convert to long format
rmt_long <- rmt %>% 
  pivot_longer(-date, names_to = "country", values_to = "remittances")

# Create tsibble
rmt_long <- rmt_long %>% 
  mutate(month_year = yearmonth(date))

ts_data <- as_tsibble(rmt_long, key = country, index = month_year)

# Inspect data
glimpse(ts_data)
```


---

# 2. Data Visualization

## 2.1 Time Series Plot
```{r ts-plot}
#| fig-width: 10
#| fig-height: 5

ts_data %>%
  filter(country == params$country) %>%
  autoplot(remittances) +
  labs(
    title = paste("Monthly Remittances from", params$country),
    subtitle = "Historical trend from July 2010 to August 2025",
    x = "Date",
    y = "Remittances (Million USD)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## 2.2 Seasonal Decomposition
```{r decomposition}
#| fig-width: 10
#| fig-height: 7

ts_data %>%
  filter(country == params$country) %>%
  model(
    STL = STL(remittances ~ trend(window = 13) + season(window = "periodic"))
  ) %>%
  components() %>%
  autoplot() +
  labs(title = paste("Seasonal Decomposition:", params$country)) +
  theme_minimal()
```

## 2.3 Seasonal Patterns
```{r seasonal-plot}
#| fig-width: 10
#| fig-height: 5

ts_data %>%
  filter(country == params$country) %>%
  gg_season(remittances, labels = "both") +
  labs(
    title = paste("Seasonal Plot:", params$country),
    y = "Remittances (Million USD)",
    x = "Month"
  )
```

## 2.4 Subseries Plot
```{r subseries-plot}
#| fig-width: 10
#| fig-height: 6

ts_data %>%
  filter(country == params$country) %>%
  gg_subseries(remittances) +
  labs(
    title = paste("Subseries Plot:", params$country),
    y = "Remittances (Million USD)"
  )
```

---

# 3. Model Diagnostics and Selection

## 3.1 Train-Test Split
```{r train-test}
# Create train and test sets
train_data <- ts_data %>%
  filter(country == params$country) %>%
  filter(month_year < yearmonth("2025 Jan"))

test_data <- ts_data %>%
  filter(country == params$country) %>%
  filter(month_year >= yearmonth("2025 Jan"))

cat("Training period:", min(train_data$month_year), "to", max(train_data$month_year), "\n")
cat("Test period:", min(test_data$month_year), "to", max(test_data$month_year), "\n")
cat("Training observations:", nrow(train_data), "\n")
cat("Test observations:", nrow(test_data), "\n")
```

## 3.2 Model Fitting
```{r model-fit}
# Fit multiple models
fit_models <- train_data %>%
  model(
    ARIMA = ARIMA(remittances),
    ETS = ETS(remittances),
    SNAIVE = SNAIVE(remittances ~ lag(12)),
    STL_ETS = decomposition_model(
      STL(remittances ~ trend(window = 13) + season(window = "periodic")),
      ETS(season_adjust ~ error("A") + trend("A") + season("N"))
    )
  )

# Generate forecasts for test period
fc_test <- fit_models %>%
  forecast(h = nrow(test_data))
```

## 3.3 Accuracy Comparison
```{r accuracy-metrics}
# Calculate accuracy metrics
accuracy_results <- fc_test %>%
  accuracy(test_data) %>%
  arrange(RMSE) %>%
  select(.model, RMSE, MAE, MAPE, MASE)

accuracy_results %>%
  gt() %>%
  tab_header(
    title = "Model Accuracy Comparison",
    subtitle = "Based on holdout test set (Jan-Aug 2025)"
  ) %>%
  fmt_number(columns = 2:5, decimals = 3) %>%
  tab_style(
    style = cell_fill(color = "#E8F5E9"),
    locations = cells_body(rows = 1)
  ) %>%
  tab_footnote(
    footnote = "Lower values indicate better performance",
    locations = cells_column_labels(columns = c(RMSE, MAE))
  )

best_model <- accuracy_results$.model[1]
```

::: {.callout-note}
## Best Model: **`r best_model`**

Based on RMSE (Root Mean Squared Error), the **`r best_model`** model performs best on the holdout test set.
:::

## 3.4 Forecast Comparison on Test Set
```{r forecast-test-plot}
#| fig-width: 10
#| fig-height: 6

fc_test %>%
  autoplot(train_data, level = NULL) +
  autolayer(test_data, remittances, color = "black", linetype = "dashed") +
  labs(
    title = "Model Comparison: Test Set Performance",
    subtitle = paste("Actual vs Predicted for", params$country),
    y = "Remittances (Million USD)",
    x = "Date"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

# 4. Final Forecast (6 Months Ahead)

## 4.1 Refit Models on Full Dataset
```{r final-models}
# Fit models on full dataset
final_models <- ts_data %>%
  filter(country == params$country) %>%
  model(
    ARIMA = ARIMA(remittances),
    ETS = ETS(remittances),
    SNAIVE = SNAIVE(remittances ~ lag(12)),
    STL_ETS = decomposition_model(
      STL(remittances ~ trend(window = 13) + season(window = "periodic")),
      ETS(season_adjust ~ error("A") + trend("A") + season("N"))
    ),
    Ensemble = (ARIMA(remittances) + ETS(remittances) + SNAIVE(remittances ~ lag(12))) / 3
  )

# Generate 6-month forecast
final_forecast <- final_models %>%
  forecast(h = params$forecast_horizon)
```

## 4.2 Model Specifications
```{r model-specs}
# Display ARIMA specification
arima_report <- final_models %>%
  select(ARIMA) %>%
  report()

# Display ETS specification
ets_report <- final_models %>%
  select(ETS) %>%
  report()

cat("### ARIMA Model Specification\n")
print(arima_report)

cat("\n### ETS Model Specification\n")
print(ets_report)
```

## 4.3 Forecast Visualization
```{r final-forecast-plot}
#| fig-width: 12
#| fig-height: 7

final_forecast %>%
  autoplot(
    ts_data %>% filter(country == params$country),
    level = c(80, 95)
  ) +
  labs(
    title = paste("6-Month Remittances Forecast:", params$country),
    subtitle = "Multiple models with 80% and 95% prediction intervals",
    y = "Remittances (Million USD)",
    x = "Date"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )
```

## 4.4 Best Model Forecast (Detailed)
```{r best-model-forecast}
#| fig-width: 10
#| fig-height: 6

# Focus on best performing model
best_forecast <- final_forecast %>%
  filter(.model == best_model)

best_forecast %>%
  autoplot(
    ts_data %>% filter(country == params$country) %>% filter(year(month_year) >= 2020),
    level = c(80, 95)
  ) +
  labs(
    title = paste("Best Model Forecast:", best_model),
    subtitle = paste("6-month ahead forecast for", params$country, "with confidence intervals"),
    y = "Remittances (Million USD)",
    x = "Date"
  ) +
  theme_minimal(base_size = 12)
```

## 4.5 Forecast Tables

### All Models
```{r forecast-table-all}
final_forecast %>%
  as_tibble() %>%
  select(month_year, .model, .mean) %>%
  pivot_wider(names_from = .model, values_from = .mean) %>%
  gt() %>%
  tab_header(
    title = paste("6-Month Forecast Comparison:", params$country),
    subtitle = "Point forecasts (mean) in Million USD"
  ) %>%
  fmt_number(columns = 2:6, decimals = 2) %>%
  tab_style(
    style = cell_fill(color = "#E8F5E9"),
    locations = cells_body(columns = all_of(best_model))
  )
```

### Best Model with Prediction Intervals
```{r forecast-table-best}
best_forecast %>%
  hilo(level = c(80, 95)) %>%
  unpack_hilo(c(`80%`, `95%`)) %>%
  select(month_year, .mean, `80%_lower`, `80%_upper`, `95%_lower`, `95%_upper`) %>%
  gt() %>%
  tab_header(
    title = paste("Best Model Forecast Details:", best_model),
    subtitle = paste(params$country, "- Point estimates with prediction intervals")
  ) %>%
  fmt_number(columns = 2:6, decimals = 2) %>%
  cols_label(
    month_year = "Month",
    .mean = "Forecast",
    `80%_lower` = "Lower",
    `80%_upper` = "Upper",
    `95%_lower` = "Lower",
    `95%_upper` = "Upper"
  ) %>%
  tab_spanner(
    label = "80% Prediction Interval",
    columns = c(`80%_lower`, `80%_upper`)
  ) %>%
  tab_spanner(
    label = "95% Prediction Interval",
    columns = c(`95%_lower`, `95%_upper`)
  )
```

---

# 5. Model Diagnostics

## 5.1 Residual Analysis
```{r residuals}
#| fig-width: 12
#| fig-height: 8

final_models %>%
  select(all_of(best_model)) %>%
  gg_tsresiduals() +
  labs(title = paste("Residual Diagnostics:", best_model))
```

## 5.2 Ljung-Box Test
```{r ljung-box}
# Perform Ljung-Box test
lb_test <- final_models %>%
  select(all_of(best_model)) %>%
  augment() %>%
  features(.innov, ljung_box, lag = 24, dof = 0)

lb_test %>%
  gt() %>%
  tab_header(
    title = "Ljung-Box Test for Residual Autocorrelation",
    subtitle = paste("Testing", best_model, "residuals")
  ) %>%
  fmt_number(columns = c(lb_stat, lb_pvalue), decimals = 4) %>%
  tab_footnote(
    footnote = "p-value > 0.05 indicates residuals resemble white noise (good)",
    locations = cells_column_labels(columns = lb_pvalue)
  )

if(lb_test$lb_pvalue > 0.05) {
  cat("\n✓ Residuals appear to be white noise (p-value =", round(lb_test$lb_pvalue, 4), ")\n")
} else {
  cat("\n⚠ Warning: Residuals may have remaining autocorrelation (p-value =", round(lb_test$lb_pvalue, 4), ")\n")
}
```

---

# 6. Forecast Summary
```{r summary-stats-forecast}
forecast_summary <- best_forecast %>%
  as_tibble() %>%
  summarise(
    Model = best_model,
    Forecast_Period = paste(min(month_year), "to", max(month_year)),
    Mean_Forecast = mean(.mean),
    Min_Forecast = min(.mean),
    Max_Forecast = max(.mean),
    Forecast_Range = Max_Forecast - Min_Forecast
  )

forecast_summary %>%
  gt() %>%
  tab_header(
    title = "Forecast Summary Statistics",
    subtitle = paste("6-month outlook for", params$country)
  ) %>%
  fmt_number(columns = 3:6, decimals = 2)
```

## Key Insights
```{r insights}
# Calculate growth rates
recent_actual <- ts_data %>%
  filter(country == params$country) %>%
  tail(12) %>%
  pull(remittances) %>%
  mean()

forecast_avg <- best_forecast %>%
  as_tibble() %>%
  pull(.mean) %>%
  mean()

pct_change <- (forecast_avg - recent_actual) / recent_actual * 100

# Trend analysis
trend_direction <- if(pct_change > 2) {
  "increasing"
} else if(pct_change < -2) {
  "decreasing"
} else {
  "stable"
}
```

::: {.callout-important}
## Forecast Highlights

- **Best performing model**: `r best_model`
- **Average forecast**: $`r round(forecast_avg, 2)` million USD
- **Expected trend**: `r trend_direction` (`r round(pct_change, 1)`% vs. recent 12-month average)
- **Forecast range**: $`r round(min(best_forecast$.mean), 2)` - $`r round(max(best_forecast$.mean), 2)` million USD
:::

---

# 7. Recommendations

Based on the analysis:

1. **Primary Model**: Use **`r best_model`** for official forecasts (lowest RMSE)

2. **Monitoring**: Track actual remittances monthly and refit models quarterly

3. **Uncertainty**: Consider the 80% prediction interval for planning purposes

4. **Seasonality**: Account for seasonal patterns observed in the data

5. **Risk Factors**: 
   - Exchange rate fluctuations
   - Economic conditions in source countries
   - Policy changes affecting remittance flows
   - Global economic conditions

---

# Appendix: Technical Details

## Data Source
- **File**: worker_remittances25.csv
- **Period**: July 2010 - August 2025
- **Frequency**: Monthly
- **Unit**: Million USD

## Models Used

1. **ARIMA**: Automatic ARIMA with seasonal components
2. **ETS**: Exponential smoothing state space model
3. **SNAIVE**: Seasonal naive with 12-month lag
4. **STL+ETS**: STL decomposition with ETS on seasonally adjusted series
5. **Ensemble**: Equal-weighted average of ARIMA, ETS, and SNAIVE

## Software
- R version 4.x
- Packages: fable, fpp3, tsibble, ggplot2

## Report Information
- **Generated**: `r Sys.Date()`
- **Forecast Horizon**: `r params$forecast_horizon` months
- **Country**: `r params$country`

---

*For questions or updates, please contact the forecasting team.*