---
title: "When Technology Outruns the Forecast: Experience Curves and Pakistan's Solar Price Collapse"
subtitle: "Wright's Law vs Moore's Law, the identification trap, and scenario-based price forecasting for policy"
author: "Prof. Dr. Zahid Asghar"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
    embed-resources: true
    include-in-header:
      text: |
        <style>
        h1, h2, h3 { color: #2C5F2D; }
        h1 { border-bottom: 2px solid #8B7355; padding-bottom: 6px; }
        a { color: #8B7355; }
        .callout-important .callout-title { background-color: #2C5F2D22; }
        body { font-family: Georgia, 'Gill Sans MT', Calibri, sans-serif; }
        </style>
params:
  data_path: "data/experiencecurveDataset_maria.csv"
execute:
  warning: false
  message: false
---

## Why technological forecasting is different

Macroeconomic forecasting treats price as a stochastic process: the trend is
"drift," a nuisance parameter estimated and extrapolated. Technological
forecasting treats price as a *function of accumulated deployment*. Wright's
Law (Wright, 1936) states that unit cost falls a fixed percentage with every
doubling of cumulative production; Moore's Law-type formulations posit a fixed
exponential decline in calendar time. The distinction is not academic. It is
the reason the IEA under-forecast solar deployment and over-forecast solar
costs in essentially every World Energy Outlook for two decades, and it is the
reason the ARIMA models of the companion analysis lost, out of sample, to a
naive random walk: a statistical model has no concept of a technology whose
cost declines *because* it is being deployed.

This note estimates the experience-curve relationship for Pakistan's solar
panel prices (January 2017 - May 2026, n = 113), subjects it to the
econometric discipline it is usually spared, and converts it into a
scenario-based five-year price forecast. Two findings complicate the naive
story in instructive ways, and both are treated as central results: the
Wright/Moore identification trap, and the reduced-form (imported) nature of
the local learning curve.

```{r}
#| label: setup
pkgs <- c("forecast", "ggplot2", "scales", "knitr")
invisible(lapply(pkgs, function(x) if (!requireNamespace(x, quietly = TRUE))
  install.packages(x, repos = "https://cran.r-project.org")))
library(forecast); library(ggplot2); library(scales); library(knitr)

brand_green <- "#2C5F2D"; brand_brown <- "#8B7355"
```

## Data

```{r}
#| label: load
df <- read.csv(params$data_path, stringsAsFactors = FALSE)
df$Date <- as.Date(df$Date, format = "%d/%m/%Y")
df <- df[order(df$Date), ]
stopifnot(!anyNA(df[, 4:6]))

p  <- df$log_commodity_price          # Pakistan panel price (log)
q  <- df$log_cumulative_capacity_mw   # Pakistan cumulative capacity (log MW)
ps <- df$log_polysilicon_usd          # global polysilicon price (log)
Q  <- exp(q); n <- nrow(df); t <- 1:n

kable(head(df, 6), caption = "First six observations")
```

```{r}
#| label: additions
#| fig-cap: "Annual capacity additions (MW): the diffusion S-curve in action"
annual_add <- tapply(Q, format(df$Date, "%Y"), max) -
              tapply(Q, format(df$Date, "%Y"), min)
add_df <- data.frame(Year = names(annual_add), MW = as.numeric(annual_add))
ggplot(add_df, aes(Year, MW)) +
  geom_col(fill = brand_green) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "MW added") + theme_minimal()
```

Cumulative capacity grew from `r round(Q[1])` MW to
`r comma(round(Q[n]))` MW - `r round((q[n]-q[1])/log(2), 1)` doublings in
nine years. Annual additions exploded from roughly 690 MW in 2017 to about
15 GW in 2024, timed exactly with the global panel price collapse, before
decelerating visibly in 2025-2026 as buy-back reform and tax uncertainty
arrived. Note for later: the deceleration matters for the scenario ceilings
in @sec-scenarios.

## Wright's Law vs Moore's Law {#sec-wright-moore}

```{r}
#| label: wright-plot
#| fig-cap: "The experience-curve view: log price against log cumulative capacity"
ggplot(df, aes(log_cumulative_capacity_mw, log_commodity_price)) +
  geom_point(color = brand_green, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = brand_brown) +
  labs(x = "log cumulative capacity (MW)", y = "log panel price") +
  theme_minimal()
```

```{r}
#| label: wright-moore
m_wright <- lm(p ~ q)      # price ~ cumulative experience
m_moore  <- lm(p ~ t)      # price ~ calendar time
b_w <- -coef(m_wright)["q"]

comp <- data.frame(
  Model = c("Wright (experience)", "Moore (time trend)"),
  Slope = c(round(coef(m_wright)["q"], 4), round(coef(m_moore)["t"], 5)),
  `R2`  = c(round(summary(m_wright)$r.squared, 3),
            round(summary(m_moore)$r.squared, 3)),
  Implied = c(paste0(percent(1 - 2^(-b_w), 0.1), " per doubling"),
              paste0(percent(1 - exp(12 * coef(m_moore)["t"]), 0.1), " per year"))
)
kable(comp, caption = "Wright vs Moore on the same series")
```

```{r}
#| label: identification
cat("cor(log capacity, time) =", round(cor(q, t), 4), "\n")
cat("R2 of q ~ t (exponential-growth check):",
    round(summary(lm(q ~ t))$r.squared, 3), "\n")
```

::: {.callout-important}
## The identification trap
The naive Wright regression implies a learning rate of
`r percent(1 - 2^(-b_w), 0.1)` per doubling - deceptively close to the
canonical ~20% global solar figure. But the Moore time-trend fits this sample
*slightly better*, and log-capacity correlates with calendar time at
`r round(cor(q, t), 3)`. This is the Nagy et al. (2013) result live in our own
data: when deployment grows exponentially, experience and time are
observationally near-equivalent, and no regression on a single national series
can establish that deployment *causes* cost decline. Any policy argument built
on "our deployment drives learning" is unidentified here.
:::

## Econometric guards {#sec-guards}

Experience curves are usually estimated as static log-log regressions on two
trending series - exactly the setting in which spurious fit flourishes. Three
guards are applied: first differences, cointegration testing with correct
critical values, and a dynamic OLS estimator for the long-run elasticities.

### First differences

```{r}
#| label: first-diff
fd  <- lm(diff(p) ~ diff(q))
fd2 <- lm(diff(p) ~ diff(q) + diff(ps))
kable(round(summary(fd2)$coefficients, 4),
      caption = "First-difference regression: monthly price changes")
```

Month to month, capacity growth predicts nothing about price changes (the
Wright slope collapses to zero, t ≈ 0.3). This is not a failure of the data -
it is the economically correct result. Pakistan holds roughly 2-3% of global
cumulative solar capacity; the learning happens in Chinese factories, and
Pakistan *imports* the learning curve. The local price-capacity relationship
also runs substantially in reverse: global price declines trigger Pakistani
adoption surges (the 2023-24 additions boom). What looks like a learning curve
is, in the short run, a diffusion curve wearing its clothes.

### Cointegration with correct critical values

```{r}
#| label: cointegration
eg2 <- lm(p ~ q)          # bivariate
eg3 <- lm(p ~ q + ps)     # trivariate
tau <- function(res) {
  r <- res; d <- diff(r); l <- r[-length(r)]
  summary(lm(d ~ l - 1))$coefficients[1, "t value"]
}
cat("EG tau, p ~ q      :", round(tau(residuals(eg2)), 3),
    " (5% MacKinnon crit ~ -3.37)\n")
cat("EG tau, p ~ q + ps :", round(tau(residuals(eg3)), 3),
    " (5% MacKinnon crit ~ -3.80)\n")
```

Price and capacity alone are not cointegrated. Adding polysilicon moves the
statistic to the borderline of its (correct, MacKinnon) critical value: the
long-run anchor, to the extent one exists, is *joint* - global technological
progress proxied by cumulative deployment, plus the upstream input cost.

### DOLS long-run elasticities and the error-correction structure

```{r}
#| label: dols
k <- 2
lead_lag <- function(x, j) {
  if (j > 0) c(x[-(1:j)], rep(NA, j)) else if (j < 0)
    c(rep(NA, -j), x[1:(length(x) + j)]) else x
}
dq <- c(NA, diff(q)); dps <- c(NA, diff(ps))
D  <- data.frame(p = p, q = q, ps = ps)
for (j in -k:k) {
  D[[paste0("dq",  j)]] <- lead_lag(dq,  j)
  D[[paste0("dps", j)]] <- lead_lag(dps, j)
}
dols <- lm(p ~ ., data = D)
cd <- summary(dols)$coefficients
b_q <- cd["q", "Estimate"]; b_ps <- cd["ps", "Estimate"]
kable(round(cd[c("q", "ps"), ], 4),
      caption = "DOLS (Stock-Watson) long-run elasticities, leads/lags k = 2")
cat("DOLS-implied learning rate:", percent(1 - 2^(b_q), 0.1), "per doubling\n")
```

```{r}
#| label: ecm
r3  <- residuals(eg3)
ecm <- lm(diff(p) ~ r3[-n] + diff(q) + diff(ps))
kable(round(summary(ecm)$coefficients, 4),
      caption = "Error-correction model for monthly price changes")
lambda <- coef(ecm)["r3[-n]"]
cat("Adjustment speed:", round(lambda, 3), "per month | half-life:",
    round(log(0.5) / log(1 + lambda), 1), "months\n")
cat("Current disequilibrium: price is",
    percent(exp(tail(r3, 1)) - 1, 0.1), "above the long-run relation\n")
```

::: {.callout-important}
## The disequilibrium premium
As of May 2026 the panel price sits roughly
`r percent(exp(tail(r3, 1)) - 1, 0.1)` **above** the estimated long-run
relation - the footprint of the 2026 GST scare and the polysilicon rebound at
the end of the sample. The error-correction term (significant, half-life about
`r round(log(0.5) / log(1 + lambda), 1)` months) implies steady downward
reversion pressure if policy uncertainty clears. This is a falsifiable
near-term prediction that no ARIMA can generate, because a drift model has no
concept of "above equilibrium." It is also a direct measure of what policy
*signaling alone* costs consumers: the 18% GST never passed, yet its shadow is
in the price.
:::

## Decomposing the 2017-2026 decline

```{r}
#| label: decomposition
dq_tot <- q[n] - q[1]; dps_tot <- ps[n] - ps[1]; dp_tot <- p[n] - p[1]
c_q  <- b_q  * dq_tot
c_ps <- b_ps * dps_tot
dec <- data.frame(
  Component = c("Capacity / technological progress (proxy)",
                "Polysilicon input cost",
                "Unexplained residual (tax, tariff, FX, margins)"),
  `Log contribution` = round(c(c_q, c_ps, dp_tot - c_q - c_ps), 3),
  `Share of fitted decline` = c(percent(c_q / (c_q + c_ps), 1),
                                percent(c_ps / (c_q + c_ps), 1), "-")
)
kable(dec, caption = "Decomposition of the actual log price change (Jan 2017 - May 2026)")
```

Roughly `r percent(c_q / (c_q + c_ps), 1)` of the fitted decline tracks the
deployment/learning proxy and `r percent(c_ps / (c_q + c_ps), 1)` tracks
polysilicon. The sizeable positive residual is domestic: it is the wedge that
GST episodes, import tariffs, rupee depreciation, and dealer margins drive
between the world module price and what a Pakistani buyer actually pays.

## Uncertainty in the learning rate

```{r}
#| label: bootstrap
set.seed(42)
B <- 2000; bl <- 12; nb <- ceiling(n / bl)
sl <- numeric(B)
for (i in 1:B) {
  starts <- sample(1:(n - bl + 1), nb, replace = TRUE)
  idx <- unlist(lapply(starts, function(s) s:(s + bl - 1)))[1:n]
  sl[i] <- coef(lm(p[idx] ~ q[idx] + ps[idx]))[2]
}
lr_ci <- 1 - 2^(quantile(sl, c(0.95, 0.5, 0.05)))
cat("Learning rate 90% interval:", percent(lr_ci[1], 0.1), "to",
    percent(lr_ci[3], 0.1), "| median", percent(lr_ci[2], 0.1), "\n")
```

A twelve-month block bootstrap puts the learning rate between roughly
`r percent(lr_ci[1], 0.1)` and `r percent(lr_ci[3], 0.1)` per doubling.
Carrying this parameter uncertainty into the forecast - rather than reporting
a single curve - is what separates the probabilistic experience-curve
tradition (Way et al., 2022) from both deterministic "law" extrapolation and
falsely-narrow ARIMA intervals.

## Scenario engine: capacity paths to price paths {#sec-scenarios}

```{r}
#| label: scenarios
fit_logistic <- function(K) {
  if (any(Q >= K)) return(NULL)
  z <- log(Q / (K - Q))
  f <- lm(z ~ t)
  list(K = K, a = coef(f)[1], b = coef(f)[2],
       ss = sum((Q - K / (1 + exp(-fitted(f))))^2))
}
grid <- Filter(Negate(is.null), lapply(seq(60000, 400000, 5000), fit_logistic))
best <- grid[[which.min(sapply(grid, `[[`, "ss"))]]
cat("Best-fit logistic saturation ceiling K =", comma(best$K), "MW\n")

H <- 60; t_f <- (n + 1):(n + H)
cap_path <- function(K, a, b) K / (1 + exp(-(a + b * t_f)))
scen <- list(
  baseline    = { f <- fit_logistic(best$K);  cap_path(f$K, f$a, f$b) },
  constrained = { f <- fit_logistic(100000);  cap_path(f$K, f$a, f$b) },
  accelerated = { f <- fit_logistic(250000);  cap_path(f$K, f$a, f$b) }
)

ps_f <- tryCatch(
  as.numeric(forecast(auto.arima(ts(ps, frequency = 12)), h = H)$mean),
  error = function(e) rep(ps[n], H))

a0 <- coef(eg3)[1]
forecast_price <- function(cap, b_q_use = b_q) {
  pf <- numeric(H); pc <- p[n]
  for (h in 1:H) {
    p_star <- a0 + b_q_use * log(cap[h]) + b_ps * ps_f[h]
    pc <- pc + lambda * (pc - p_star)
    pf[h] <- pc
  }
  pf
}
paths <- sapply(scen, forecast_price)

fan <- sapply(sample(sl, 500), function(bq) forecast_price(scen$baseline, bq))
fan_lo <- apply(fan, 1, quantile, 0.05)
fan_hi <- apply(fan, 1, quantile, 0.95)
```

The engine has three moving parts: (i) a logistic diffusion curve fitted to
cumulative capacity, with the saturation ceiling K chosen by grid search
(best fit near `r comma(best$K)` MW) and bracketed by a policy-constrained
scenario (K = 100 GW: net-billing and tax friction bite) and an accelerated
scenario (K = 250 GW: batteries unlock a second adoption wave); (ii) an
auxiliary ARIMA projection of polysilicon; (iii) the error-correction
forecaster, in which the price converges at rate
`r round(abs(lambda), 3)`/month toward a *moving* equilibrium defined by the
DOLS elasticities. The forecast therefore respects both the current
disequilibrium premium and the falling technological anchor.

```{r}
#| label: fan-chart
#| fig-cap: "Five-year forecasts: technological (scenario fan) vs statistical benchmarks"
#| fig-height: 5.5
arima_fc <- tryCatch(
  as.numeric(forecast(auto.arima(ts(p, frequency = 12)), h = H)$mean),
  error = function(e) rep(NA, H))
naive_fc <- rep(p[n], H)

dates_f <- seq(df$Date[n] + 31, by = "month", length.out = H)
plot_df <- rbind(
  data.frame(Date = df$Date, value = p, series = "Actual"),
  data.frame(Date = dates_f, value = paths[, "baseline"],    series = "Wright-ECM baseline (K=150GW)"),
  data.frame(Date = dates_f, value = paths[, "constrained"], series = "Policy-constrained (K=100GW)"),
  data.frame(Date = dates_f, value = paths[, "accelerated"], series = "Accelerated (K=250GW)"),
  data.frame(Date = dates_f, value = arima_fc,               series = "ARIMA drift"),
  data.frame(Date = dates_f, value = naive_fc,               series = "Naive")
)
ribbon_df <- data.frame(Date = dates_f, lo = fan_lo, hi = fan_hi)
ggplot(plot_df, aes(Date, value, color = series)) +
  geom_ribbon(data = ribbon_df, aes(x = Date, ymin = lo, ymax = hi),
              inherit.aes = FALSE, alpha = 0.15, fill = brand_green) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = c(
    "Actual" = "black",
    "Wright-ECM baseline (K=150GW)" = brand_green,
    "Policy-constrained (K=100GW)" = brand_brown,
    "Accelerated (K=250GW)" = "#4C9A2A",
    "ARIMA drift" = "steelblue", "Naive" = "grey55")) +
  labs(y = "log panel price", x = NULL, color = NULL,
       subtitle = "Ribbon: 90% learning-rate uncertainty on the baseline") +
  theme_minimal() + theme(legend.position = "bottom")
```

```{r}
#| label: forecast-table
fc_tbl <- data.frame(
  Horizon      = c("12 months", "36 months", "60 months"),
  Baseline     = round(paths[c(12, 36, 60), "baseline"], 3),
  Constrained  = round(paths[c(12, 36, 60), "constrained"], 3),
  Accelerated  = round(paths[c(12, 36, 60), "accelerated"], 3),
  ARIMA        = round(arima_fc[c(12, 36, 60)], 3),
  Naive        = round(naive_fc[c(12, 36, 60)], 3)
)
kable(fc_tbl, caption = "Forecast comparison (log price)")
cat("Baseline implied cumulative price change at 60 months:",
    percent(exp(paths[60, "baseline"] - p[n]) - 1, 0.1), "\n")
```

The baseline implies a further decline of roughly
`r percent(1 - exp(paths[60, "baseline"] - p[n]), 0.1)` over five years, but
the honest headline is the width of the fan: at 60 months the 90% band spans
about `r round(fan_lo[60], 2)` to `r round(fan_hi[60], 2)` in log terms. The
scenarios diverge slowly at 12 months and widely at 60 - technological
forecasts earn their keep precisely at the horizons where statistical models
extrapolate blind.

## What this means for policy

::: {.callout-important}
## Policy relevance

- **Planning institutions are structurally exposed.** IGCEP-style generation
  planning and DISCO revenue models built on ARIMA-class or static cost
  assumptions will keep being ambushed: Pakistan's ~56 GW of largely
  behind-the-meter capacity arrived because official projections treated cost
  declines as drift rather than as a function of deployment. The February 2026
  net-billing rush is the institutional aftershock of a forecasting failure.
- **Suppressing adoption does not stop the cost decline - it shifts its
  composition.** An 18% GST or an Rs 11/unit buy-back pushes households toward
  self-consumption plus storage; batteries sit on their own ~25-30% learning
  curve and will hit the grid as a second, larger surprise on a predictable
  schedule.
- **Policy signaling has a measurable price.** The
  `r percent(exp(tail(r3, 1)) - 1, 0.1)` disequilibrium premium at the end of
  the sample is the cost of GST *uncertainty alone* - the tax never passed,
  yet its shadow is in the price paid by every buyer in early 2026.
- **The right planning input is a scenario range, not a point forecast.**
  Tariff design, cross-subsidy recovery, and IPP capacity-payment exposure
  should be stress-tested against the fan above, in the spirit of Way et al.
  (2022): on experience-curve evidence, the fast-transition path is the
  cheap path, and models that assume static technology costs compound against
  the state.
:::

## Key references

Nagy, B., Farmer, J. D., Bui, Q. M., & Trancik, J. E. (2013). Statistical
basis for predicting technological progress. *PLoS ONE*, 8(2), e52669.

Way, R., Ives, M. C., Mealy, P., & Farmer, J. D. (2022). Empirically grounded
technology forecasts and the energy transition. *Joule*, 6(9), 2057-2082.

Wright, T. P. (1936). Factors affecting the cost of airplanes. *Journal of the
Aeronautical Sciences*, 3(4), 122-128.

Kavlak, G., McNerney, J., & Trancik, J. E. (2018). Evaluating the causes of
cost reduction in photovoltaic modules. *Energy Policy*, 123, 700-710.

## Reproducibility

```{r}
#| label: session-info
#| code-fold: true
sessionInfo()
```
