---
title: "Spillover Effect NIFTY50 and SP500"
format: docx
execute: 
  freeze: auto
  warning: false
---






## Load libraries

```{r}
library(tidyverse)

library(xts)

library(zoo)


library(janitor)
library(tseries)

library(Spillover)
library(gt)
library(urca)
```


## Load data

```{r}
dyn_df <- read_csv('data/indo_us.csv') 


dyn_df |> glimpse()  # inspect data


dyn_df$date <- dmy(dyn_df$Date) #convert Date from string to date format

dyndf <- dyn_df |> dplyr::select(date, NIFTY50,SP500) |> clean_names() #all names in short alphabets

```


## Time Plot  

To inspect time series data, first step is always to plot and see how it looks like.
```{r}
# Time plots to see idea how indices move 

dyndf |> 
  pivot_longer(cols = c(nifty50, sp500), names_to = "Variable", values_to = "Value") |> 
ggplot()+ aes(x = date, y = log(Value), color = Variable) +
  geom_line() +
  labs(title = "Multiple Line Graph", x = "Date", y = "Y-Axis Label", color = "Variable") +
  theme_minimal()
```


## Descriptive Statistics

```{r}
library(modelsummary)

dyndf  |> datasummary_skim()

# Or another way to have summary of data



```


Codes for this summary are not displayed in the output


```{r}
#|echo: false
# Calculate statistics for nifty50
mean_nifty50 <- mean(dyndf$nifty50)
sd_nifty50 <- sd(dyndf$nifty50)
var_nifty50 <- var(dyndf$nifty50)
max_nifty50 <- max(dyndf$nifty50)
min_nifty50 <- min(dyndf$nifty50)

# Calculate statistics for sp500
mean_sp500 <- mean(dyndf$sp500)
sd_sp500 <- sd(dyndf$sp500)
var_sp500 <- var(dyndf$sp500)
max_sp500 <- max(dyndf$sp500)
min_sp500 <- min(dyndf$sp500)

# Perform Jarque-Bera test for nifty50
jb_test_nifty50 <- jarque.bera.test(dyndf$nifty50)
# Perform Jarque-Bera test for sp500
jb_test_sp500 <- jarque.bera.test(dyndf$sp500)

# Perform ADF test for nfity50
adf_test_nfity50 <- ur.df(dyndf$nifty50, lags = 1)

# Perform ADF test for ts2
adf_test_sp500 <- ur.df(dyndf$sp500, lags = 1)

# Create a data frame for the results
results <- data.frame(
  Variable = c("nifty50", "sp500"),
  Mean = c(mean_nifty50, mean_sp500),
  SD = c(sd_nifty50, sd_sp500),
  Variance = c(var_nifty50, var_sp500),
  Max = c(max_nifty50, max_sp500),
  Min = c(min_nifty50, min_sp500),
  JB_P_Value = c(2.43975e-07, 2.129131e-06 ),
  ADF_Statistic = c(1.53, 1.063)
)

results

library(tidyr)
library(gt)

results %>%
  gather(Statistic, Value, -Variable) |> gt() |> fmt_number(decimals = 3)


```





##  Variance decomposition using VAR 

VAR(2) and 10-day-ahead variance decomposition.


```{r}
dyndf %>% 
  dplyr::select(-date) %>% 
  VAR(p=2) %>% 
  G.spillover(standardized = FALSE) %>% 
  round(2)
```




## Now, let’s move on, and get the dynamic total spillover index:


```{r}

# total_dynamic <- total.dynamic.spillover(as.zoo(dyndf[,-1]),
#                                          width = 20, 
#                                          index="generalized",
#                                          p=2)

load("data/total_dynamic.RData")
tibble(Date= dyndf$date[-c(1:19)],  index = total_dynamic) %>% 
   ggplot(aes(x=Date, y=index)) +
  geom_line()+
  labs(caption = "Fig 2. Total volatility spillovers for both indices")+
  theme(plot.caption = element_text(hjust = 0))

```


## 

#### Directional volatility spillovers, FROM four asset classes
#### In order to get Figure 3 from Diebold and Yilmaz (2012), we proceed as follows:


```{r}
# dy_results <- dynamic.spillover(data=dyndf, 
#                                 width=20, 
#                                 p=2) 

load("data/dy_results.RData")
pp_from <- plotdy(data=dy_results, direction = "from") 


pp_from +
  labs(caption = "Fig 3. Directional volatility spillovers, FROM four asset classes.")+
  theme(plot.caption = element_text(hjust = 0.5))  
pp_to <- plotdy(dy_results, direction = "to")  


# Just for customization 
pp_from +
  labs(caption = "Fig. 4. Directional volatility spillovers, TO four asset classes.")+
  theme(plot.caption = element_text(hjust = 0.5)) 

```


## 


```{r}
pp_net <- plotdy(dy_results, direction = "net")  

pp_net +
  labs(caption = "Fig. 5. Net volatility spillovers, four asset classes.")+
  theme(plot.caption = element_text(hjust = 0.5))

```


## Net pairwise volatility spillovers.

```{r}
pp_netpairwise <- plotdy(dy_results, direction = "net_pairwise")  

pp_netpairwise +
  labs(caption = "Fig. 6. Net pairwise volatility spillovers.")+
  theme(plot.caption = element_text(hjust = 0.5))  

pp_from_to_pairwise <- plotdy(dy_results, direction = "from_to_pairwise")

```




