---
title: 'Understanding econometric modeling with simple <br> Consumption Function for Pakistan  '
subtitle: "Model Selection, Multicollinearity, G2S <br> Course: Economic Forecasting and Modeling"
author: '*Zahid Asghar *'
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  revealjs:
    theme: [default, custom.scss]
    logo: ""
    footer: "[Zahid Asghar](https://zahidasghar.com)"
    height: 900
    width: 1600
    chalkboard: true
    title-slide-attributes: 
      data-background-color: "#447099"
overview: true
execute:
  echo: false
  warning: false
  freeze: auto
resources:
  - slides.pdf
---

```{r include=FALSE}
library(tidyverse)
library(httr)
clrs <- MetBrewer::met.brewer(name = "Java")
clrs_lt <- colorspace::lighten(clrs, 0.9)
knitr::opts_chunk$set(fig.retina = 3, collapse = TRUE)
options(digits = 3, width = 75)
```

## Keynesian Consumption Function {background-color="`r clrs[1]`"}

-   Within time series models, the simplest type is the static model.
-   all the action takes place within one time period,
-   Keynesian consumption function says that current consumption depends only on current income, so that $C_t = f(Y_t)$ : $C_t =\alpha + \beta Y_t + \epsilon_t$

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(dynlm)
library(haven)
library(fpp3)

```

## Load data {background-color="`r clrs_lt[1]`"}

```{r}
#| warning: false
#| message: false
cons_inc<-read_dta("Consumption_Income_Pakistan.dta")
library(tseries)
cons_inc$year<-as_date(cons_inc$year)
cons_inc<-cons_inc %>% mutate(Trend=1:52)

```

### Plot of the data {background-color="`r clrs_lt[1]`"}

```{r}
#| echo: false
#| warning: false
#| message: false
#cons_inc<-ts(con_inc,start = 1960, end=2019)

#cons_inc<-as_tibble(con_inc)
ggplot(cons_inc)+aes(x=year)+geom_line(aes(y = C), color = "darkred",size=0.8) + 
  geom_line(aes(y = Y), color="steelblue", linetype="twodash",size=0.8)+labs(x="Income per capita",y="Consumption per captia",title = "Consumption Income pattern over time for Pakistan")+theme_minimal()
```

## 

The causal hypothesis is GDP $\Rightarrow$ Con.  $C = \alpha + \beta Y+\epsilon$. Running this regression on the data leads to the following results:


```{r, echo=FALSE, warning=FALSE,message=FALSE}
#| warning: false
#| message: false
library(kableExtra)
library(stargazer)
library(broom)
library(moderndive)
library(gtsummary)
library(gtExtras)
con_mod<-lm(C~Y,data = cons_inc)
regression_points <- get_regression_points(con_mod)
regression_points %>% kable()
```

## 

::: panel-tabset
### tidy model

```{r}
library(ggfortify)
library(modelsummary)
autoplot(con_mod)
cons_inc<-cons_inc|>mutate(resid=residuals(con_mod))
tidy(con_mod)
```

### model summary {.scroll}

```{r}
modelsummary(con_mod)
#get_regression_table(con_mod)
#library(modelsummary)
#modelsummary(con_mod)
```
:::
## 
$R^2$ of 0.97, $t-stat=\frac{0.785}{0.020}=39$ , $p-value$ of $0.000$ means we can reject the null hypothesis that the true coefficient is zero,

## Residual diagnostic

::: panel-tabset
### Residuals-plot

```{r}
#acf(resid)
#pacf(resid)
#adf.test(resid)
ggplot(cons_inc)+aes(x=year, y=resid)+geom_col()
```

### Diagnostic tests

```{r, echo=FALSE,warning=FALSE,message=FALSE}
library(ggResidpanel)
resid_panel(con_mod,plots = 'default',smoother = TRUE)

```
:::

## Deterministic trend

```{r, echo=FALSE}
library(gridExtra)
g11<-ggplot(cons_inc, aes(x = Trend, y = C)) +
  geom_point() +
  labs(x = "Time trend", y = "Consumption",
       title = "Relationship between Consumption with time trend") +  
  geom_smooth(method = "lm", se = FALSE)
tren_mod<-lm(C~Trend,data = cons_inc)
cons_inc<-cons_inc|>mutate(trend_resid=tren_mod$residuals)
g12<-ggplot(cons_inc)+aes(x=year, y=trend_resid)+geom_col()
library(patchwork)
g11+g12
```

## What to do when errors are not independent? {.scroll}

$e_t=y_t-\hat{\beta}x_t$ and $e_{t-1}=y_{t-1}-\hat{\beta}x_{t-1}$. A relationship between the two errors shows that what happens in period $t-1$ has an effect on what happens in period $t$. In other words, we have a dynamic model instead of a static model.

### STATIC VERSUS DYNAMIC MODELS

$C_t = \alpha + \beta Y_t + e_t$ : A static model.

Suppose that we conduct tests and learn that $e_t$ is correlated with $e_{t-1}$. $e_{t-1}=C_{t-1}-\alpha-\beta Y_{t-1}$. It follows that $C_{t-1}$ and $Y_{t-1}$ have an effect on current consumption.

The simplest dynamic extension of the Keynesian model is the following: $$C_t=\alpha+\beta Y_t+\gamma C_{t-1}+\delta Y_{t-1}+e_t$$

# General-to-Simple vs. Simple-to-General Modelling {background-color="`r clrs[1]`"}

## Dynamic Model {.scroll}

-   **General to Specific** approach and include upto three lags of both consumption and income.
-   No hard and fast rule : Regressors should not exceed
-   square root of observations
-   1/3 of the total number of observations

## {.scroll}
```{r}
#| warning: false
#| echo: false
#| message: false
g2s<-lm(C~Y+lag(Y,1)+lag(C,1)+lag(Y,2)+lag(C,2)+lag(Y,3)+lag(C,3),data=cons_inc)

modelsummary(g2s,estimate = c("{estimate} ({std.error}){stars}"))
```

## Re-run by exclusing non-significant variables {.small}

```{r}
#| warning: false
#| message: false

g2s1<-lm(C~Y+lag(C,1)+lag(Y,2),data=cons_inc)
dyn_inc<-cons_inc|>filter(year>1968)|>mutate(g2s1res=g2s1$residuals)|>na.omit()
g13<-ggplot(dyn_inc)+aes(x=year, y=g2s1res)+geom_col()

modelsummary(g2s1,estimate = c("{estimate} ({std.error}){stars}"))

```

## 

> ### Oh great now we have all the variables are highly significant and if residuals satisfy all standard assumptions, it will be perfect fit. So lets have a look at the residuals behavior.

## 

Lets have first look at residuals plot over time.

```{r}
#| echo: false
g13
```

## 

It seems residuals are behaving randomly. Lets apply some more tests before I finalize the model.

```{r}
#| warning: false
#| echo: false
#| message: false

resid_panel(g2s1,plots = 'default',smoother = TRUE)
```

This seems quite a good job. But what about [**theoretical interpretation**]{.purple}.

## Multicolinearity and economic rationale

::: border
How to explain that second lag of Y has an impact while first lag of Y has not?
:::

```{r}
#| echo: false
library(corrplot)
cons_inc<-cons_inc|>mutate(lagY1=lag(Y,1),lagY2=lag(Y,2),lagY3=lag(Y,3),lagC1=lag(C,1),lagC2=lag(C,2),lagC3=lag(C,3))
M<-cons_inc|>select(C,Y,lagY1,lagY2,lagY3,lagC1,lagC2,lagC3)|>na.omit()
M1<-cor(M)
round(M1,3) %>% kable()
```

## Model with first lag of Y {.scroll}

```{r}
#| echo: false

g2s2<-lm(C~Y+lag(C,1)+lag(Y,1),data=cons_inc)

modelsummary(g2s2,estimate = c("{estimate} ({std.error}){stars}"))
```

## 

> ### Oh that makes sense and current consumption is explained by current income, past year consumption and past year income and all other variables are insignificant. This model pass all diagnostic tests as well.

## 

```{r}
#| warning: false
#| echo: false
#| message: false

resid_panel(g2s2,plots = 'default',smoother = TRUE)
```

## Summary

-   Theory
-   Statistical Model
-   Data
-   Estimation Methods
