---
title: "Time Series Analysis: KSE-100 Index"
subtitle: "A Comprehensive Tutorial Following Hyndman's Approach"
author: "Your Name"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
  cache: false
  error: true
  freeze: auto
editor_options: 
  chunk_output_type: console
---

```{r setup}
#| include: false
library(tidyverse)
library(fpp3)
library(tsibble)
library(feasts)
library(forecast)
library(dygraphs)
library(xts)
library(zoo)
library(janitor)
library(knitr)
library(kableExtra)

# Set theme for ggplot
theme_set(theme_minimal(base_size = 12))
```

# Introduction {.tabset}

## Overview

Welcome to this comprehensive time series analysis tutorial using the **Karachi Stock Exchange (KSE-100) Index**. This tutorial follows the pedagogical approach of Professor Rob J. Hyndman, emphasizing:

- **Visual exploration** before modeling
- **Understanding patterns** in data
- **Systematic decomposition** of time series components
- **Forecasting** with appropriate models

### Learning Objectives

By the end of this tutorial, you will be able to:

1. Load and prepare time series data
2. Create effective visualizations for time series
3. Identify and understand time series components (trend, seasonality, cycles)
4. Decompose time series using classical and modern methods
5. Check for stationarity and transformations
6. Build forecasting models

### The KSE-100 Index

The KSE-100 Index is a stock index acting as a benchmark to compare prices on the Pakistan Stock Exchange over time. It captures the performance of the top 100 companies listed on the exchange.

## Data Description

Our dataset contains daily trading information for the KSE-100 index:

**Variables:**

- `date`: Trading date
- `open`: Opening price
- `high`: Highest price during the day
- `low`: Lowest price during the day
- `close`: Closing price
- `vol`: Trading volume
- `change_percent`: Daily percentage change

**Time Period:** The data spans multiple years of trading activity.

## Setup Requirements

```r
# Required packages
install.packages(c("tidyverse", "fpp3", "tsibble", 
                   "feasts", "forecast", "dygraphs", 
                   "xts", "zoo", "janitor"))
```

# Data Loading & Preparation {.tabset}

## Load Data

```{r load-data}
# Load the dataset
kse_raw <- read_csv("D:/RepTemplates/timeseries/docs/data/kse_2025.csv") %>% 
  clean_names()

# Display first few rows
head(kse_raw) %>%
  kable(caption = "First 6 rows of raw KSE-100 data") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Data Cleaning

```{r clean-data}

kse <- kse_raw %>%
  mutate(
    # robust date parsing
    date = mdy(trimws(date)),

    # % to numeric
    change_percent = parse_number(change_percent),

    # handle 85.44M / 190.65M / 1.2B / 930K, commas, spaces, etc.
    vol_clean = str_trim(vol),
    vol_base  = parse_number(vol_clean),  # strips commas, keeps number
    vol_mult  = case_when(
      str_detect(vol_clean, regex("M$", ignore_case = TRUE)) ~ 1e6,
      str_detect(vol_clean, regex("B$", ignore_case = TRUE)) ~ 1e9,
      str_detect(vol_clean, regex("K$", ignore_case = TRUE)) ~ 1e3,
      TRUE ~ 1
    ),
    vol = vol_base * vol_mult
  ) %>%
  select(date, open, high, low, vol, change_percent) %>%
  arrange(date) %>%
  tidyr::drop_na()

# Create tsibble object (tidy time series)
kse_tsbl <- kse %>%
  as_tsibble(index = date)

# Summary statistics
summary(kse) %>%
  kable(caption = "Summary Statistics of KSE-100 Variables") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Data Structure

```{r data-structure}
# Display structure
glimpse(kse_tsbl)

# Check for gaps in the time series
has_gaps(kse_tsbl)

# Count observations by year
kse_tsbl %>%
  mutate(year = year(date)) %>%
  count(year) %>%
  ggplot(aes(x = factor(year), y = n)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = n), vjust = -0.5) +
  labs(title = "Number of Trading Days per Year",
       x = "Year", y = "Count") +
  theme_minimal()
```

# Visual Exploration {.tabset}

## Time Plot

> "The first thing to do in any data analysis task is to plot the data." - Rob J. Hyndman

```{r time-plot}
# Static time plot
kse_tsbl %>%
  autoplot(open) +
  labs(title = "KSE-100 Index: Closing Prices",
       y = "Closing Price",
       x = "Date") +
  theme_minimal()

# Interactive time plot
kse_xts <- xts(kse$open, order.by = kse$date)
dygraph(kse_xts, main = "KSE-100 Index (Interactive)") %>%
  dyAxis("y", label = "Close Price") %>%
  dyRangeSelector() %>%
  dyOptions(colors = "#2C3E50", strokeWidth = 1.5)
```

**Key Observations:**

- Identify overall trend direction
- Note any major jumps or crashes
- Look for periods of high volatility
- Observe general price levels across time

## Multiple Variables

```{r multiple-vars}
# Plot all price variables
kse_tsbl %>%
  select(date, open, high, low) %>%
  pivot_longer(cols = -date, names_to = "price_type", values_to = "price") %>%
  ggplot(aes(x = date, y = price, color = price_type)) +
  geom_line(alpha = 0.7) +
  labs(title = "KSE-100: All Price Types",
       y = "Price", x = "Date",
       color = "Price Type") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Daily Returns

```{r returns}
# Calculate daily returns
kse_returns <- kse_tsbl %>%
  mutate(
    daily_return = (open - lag(open)) / lag(open) * 100,
    abs_return = abs(daily_return)
  )

# Plot returns
kse_returns %>%
  ggplot(aes(x = date, y = daily_return)) +
  geom_line(color = "darkblue", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "KSE-100 Daily Returns",
       y = "Daily Return (%)",
       x = "Date") +
  theme_minimal()

# Distribution of returns
kse_returns %>%
  ggplot(aes(x = daily_return)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Distribution of Daily Returns",
       x = "Daily Return (%)",
       y = "Frequency") +
  theme_minimal()
```

## Volume Analysis

```{r volume}
# Volume over time
kse_tsbl %>%
  ggplot(aes(x = date, y = vol)) +
  geom_line(color = "darkgreen", alpha = 0.7) +
  labs(title = "KSE-100 Trading Volume",
       y = "Volume",
       x = "Date") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

# Price vs Volume relationship
kse_tsbl %>%
  ggplot(aes(x = vol, y = open)) +
  geom_point(alpha = 0.3, color = "darkblue") +
  geom_smooth(method = "loess", color = "red") +
  labs(title = "Price vs Volume Relationship",
       x = "Volume",
       y = "Close Price") +
  scale_x_continuous(labels = scales::comma) +
  theme_minimal()
```

# Seasonal Patterns {.tabset}

## Seasonal Plots

```{r seasonal-plots}
# Add time features
kse_seasonal <- kse_tsbl %>%
  mutate(
    year = year(date),
    month = month(date, label = TRUE),
    week = week(date),
    wday = wday(date, label = TRUE)
  )

# Monthly patterns
kse_seasonal %>%
  ggplot(aes(x = month, y = open, group = year, color = factor(year))) +
  geom_line(alpha = 0.6) +
  labs(title = "KSE-100: Seasonal Pattern by Month",
       x = "Month", y = "Close Price",
       color = "Year") +
  theme_minimal() +
  theme(legend.position = "right")

```

## Day of Week Effects

```{r weekday-effects}
# Day of week analysis
weekday_summary <- kse_returns %>%
  mutate(wday = wday(date, label = TRUE)) %>%
  group_by(wday) %>%
  summarise(
    avg_return = mean(daily_return, na.rm = TRUE),
    sd_return = sd(daily_return, na.rm = TRUE),
    n = n()
  )

weekday_summary %>%
  ggplot(aes(x = wday, y = avg_return)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_errorbar(aes(ymin = avg_return - sd_return, 
                    ymax = avg_return + sd_return),
                width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Average Daily Returns by Day of Week",
       x = "Day of Week",
       y = "Average Return (%)") +
  theme_minimal()

#weekday_summary %>%
 ##      col.names = c("Day", "Avg Return (%)", "Std Dev", "Count"),
   #     digits = 3) %>%
  #able_styling(bootstrap_options = c("striped", "hover"))
```

## Monthly Aggregation

```{r monthly-agg}
# Aggregate to monthly data
kse_monthly <- kse_tsbl %>%
  mutate(year_month = yearmonth(date)) %>%
  as_tibble() %>%  # Convert to tibble first to remove tsibble structure
  group_by(year_month) %>%
  summarise(
    avg_open = mean(open, na.rm = TRUE),
    max_open = max(open, na.rm = TRUE),
    min_open = min(open, na.rm = TRUE),
    total_vol = sum(vol, na.rm = TRUE),
    .groups = "drop"  # Automatically drop grouping
  ) %>%
  as_tsibble(index = year_month)

# Plot monthly average
kse_monthly %>%
  autoplot(avg_open) +
  labs(title = "KSE-100: Monthly Average Closing Price",
       y = "Average open Price",
       x = "Month") +
  theme_minimal()
```

# Decomposition {.tabset}

## Classical Decomposition

```{r classical-decomp}
# Check if we have enough data for classical decomposition
n_months <- nrow(kse_monthly)
cat("Number of months available:", n_months, "\n")

# Classical decomposition requires at least 2 complete periods (24 months)
if (n_months >= 24) {
  # Monthly data decomposition
  dcmp_classical <- kse_monthly %>%
    model(
      classical = classical_decomposition(avg_open, type = "additive")
    )
  
  # Plot components
  components(dcmp_classical) %>%
    autoplot() +
    labs(title = "Classical Decomposition: KSE-100 Monthly Average") +
    theme_minimal()
  
  # Extract components
  components(dcmp_classical) %>%
    as_tibble() %>%
    head(12) %>%
    kable(caption = "First 12 months of decomposed components",
          digits = 2) %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
} else {
  cat("⚠️ Not enough data for classical decomposition.\n")
  cat("Need at least 24 months, have", n_months, "months.\n\n")
  cat("Using a simple trend extraction instead:\n\n")
  
  # Alternative: simple moving average trend
  kse_monthly %>%
    mutate(
      trend = slider::slide_dbl(avg_open, mean, 
                                .before = 2, .after = 2, .complete = TRUE)
    ) %>%
    pivot_longer(cols = c(avg_open, trend), 
                 names_to = "component", 
                 values_to = "value") %>%
    ggplot(aes(x = year_month, y = value, color = component)) +
    geom_line(size = 1) +
    scale_color_manual(values = c("avg_open" = "gray60", "trend" = "darkblue"),
                      labels = c("avg_open" = "Original", "trend" = "Trend (5-MA)")) +
    labs(title = "KSE-100: Original Series with Trend",
         y = "Average open Price",
         x = "Month",
         color = "") +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

## STL Decomposition

> "STL (Seasonal and Trend decomposition using Loess) is a versatile and robust method for decomposing time series." - Hyndman

```{r stl-decomp}
# STL decomposition also requires sufficient data
if (n_months >= 24) {
  # STL decomposition
  dcmp_stl <- kse_monthly %>%
    model(
      stl = STL(avg_open ~ season(window = "periodic"))
    )
  
  # Plot STL components
  components(dcmp_stl) %>%
    autoplot() +
    labs(title = "STL Decomposition: KSE-100 Monthly Average") +
    theme_minimal()
  
  # Seasonally adjusted series
  kse_monthly %>%
    autoplot(avg_open, color = "gray") +
    autolayer(components(dcmp_stl), season_adjust, color = "blue") +
    labs(title = "KSE-100: Original vs Seasonally Adjusted",
         y = "Average open Price",
         x = "Month") +
    theme_minimal()
} else {
  cat("⚠️ Not enough data for STL decomposition.\n")
  cat("Using alternative trend-cycle extraction:\n\n")
  
  # Use a simple loess smoother for trend
  kse_monthly %>%
    mutate(
      trend = predict(loess(avg_open ~ as.numeric(year_month), 
                           data = ., span = 0.3))
    ) %>%
    ggplot(aes(x = year_month)) +
    geom_line(aes(y = avg_open), color = "gray60", alpha = 0.7) +
    geom_line(aes(y = trend), color = "darkred", size = 1.2) +
    labs(title = "KSE-100: Original Series with LOESS Trend",
         y = "Average open Price",
         x = "Month") +
    theme_minimal()
}
```

## Trend Analysis

```{r trend-analysis}
# Extract and analyze trend
if (exists("dcmp_stl") && n_months >= 24) {
  trend_data <- components(dcmp_stl) %>%
    as_tibble() %>%
    select(year_month, trend)
  
  # Plot trend with confidence interval
  ggplot(trend_data, aes(x = year_month, y = trend)) +
    geom_line(color = "darkblue", size = 1) +
    geom_smooth(method = "loess", color = "red", fill = "pink", alpha = 0.3) +
    labs(title = "KSE-100 Trend Component with Smoothed Trend",
         y = "Trend",
         x = "Month") +
    theme_minimal()
  
  # Calculate trend statistics
  trend_stats <- trend_data %>%
    summarise(
      min_trend = min(trend, na.rm = TRUE),
      max_trend = max(trend, na.rm = TRUE),
      avg_trend = mean(trend, na.rm = TRUE),
      trend_range = max_trend - min_trend,
      pct_change = (max_trend - min_trend) / min_trend * 100
    )
  
  trend_stats %>%
    kable(caption = "Trend Component Statistics",
          digits = 2) %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
} else {
  # Alternative trend analysis using moving average
  kse_monthly %>%
    mutate(
      ma_3 = slider::slide_dbl(avg_open, mean, .before = 1, .after = 1, .complete = TRUE),
      ma_6 = slider::slide_dbl(avg_open, mean, .before = 2, .after = 2, .complete = TRUE)
    ) %>%
    pivot_longer(cols = c(avg_open, ma_3, ma_6),
                 names_to = "series",
                 values_to = "value") %>%
    ggplot(aes(x = year_month, y = value, color = series)) +
    geom_line(size = 1) +
    scale_color_manual(
      values = c("avg_open" = "gray50", "ma_3" = "blue", "ma_6" = "darkred"),
      labels = c("avg_open" = "Original", "ma_3" = "3-Month MA", "ma_6" = "6-Month MA")
    ) +
    labs(title = "KSE-100: Trend Extraction using Moving Averages",
         y = "Average open Price",
         x = "Month",
         color = "") +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  # Simple trend statistics
  kse_monthly %>%
    summarise(
      first_value = first(avg_open),
      last_value = last(avg_open),
      min_value = min(avg_open),
      max_value = max(avg_open),
      total_change = last_value - first_value,
      pct_change = (last_value - first_value) / first_value * 100
    ) %>%
    kable(caption = "Overall Trend Statistics",
          digits = 2) %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
}
```

# Autocorrelation {.tabset}

## ACF Analysis

```{r acf}
# Note: Stock market data has irregular gaps (weekends, holidays)
# We need to use fill_gaps() or work with the index directly

# ACF plot for closing prices
kse_tsbl %>%
  fill_gaps() %>%  # Fill gaps with NA
  ACF(open, lag_max = 50) %>%
  autoplot() +
  labs(title = "ACF: KSE-100 Closing Prices",
       subtitle = "Strong persistence indicates non-stationarity") +
  theme_minimal()

# Alternative: Use acf() from stats package with regular spacing
acf_open <- acf(kse$open, lag.max = 50, plot = FALSE)
autoplot(acf_open) +
  labs(title = "ACF: KSE-100 Closing Prices (Alternative Method)",
       x = "Lag", y = "ACF") +
  theme_minimal()

# ACF for returns
kse_returns %>%
  fill_gaps() %>%
  ACF(daily_return, lag_max = 50) %>%
  autoplot() +
  labs(title = "ACF: Daily Returns",
       subtitle = "Rapid decay suggests openr to stationarity") +
  theme_minimal()

# Alternative for returns
returns_clean <- na.omit(kse_returns$daily_return)
acf_returns <- acf(returns_clean, lag.max = 50, plot = FALSE)
autoplot(acf_returns) +
  labs(title = "ACF: Daily Returns (Alternative Method)",
       x = "Lag", y = "ACF") +
  theme_minimal()
```

**Interpretation:**

- **Closing prices**: Strong autocorrelation at all lags indicates non-stationarity
  - Values highly correlated with recent past
  - Slow decay typical of price series
  
- **Returns**: Autocorrelation quickly drops to zero (openr to stationarity)
  - Little predictability from past returns
  - Consistent with efficient market hypothesis

## PACF Analysis

```{r pacf}
# PACF plot for closing prices
kse_tsbl %>%
  fill_gaps() %>%
  PACF(open, lag_max = 50) %>%
  autoplot() +
  labs(title = "PACF: KSE-100 Closing Prices",
       subtitle = "High lag-1 value indicates strong AR component") +
  theme_minimal()

# Alternative method
pacf_open <- pacf(kse$open, lag.max = 50, plot = FALSE)
autoplot(pacf_open) +
  labs(title = "PACF: KSE-100 Closing Prices (Alternative Method)",
       x = "Lag", y = "PACF") +
  theme_minimal()

# PACF for returns
kse_returns %>%
  fill_gaps() %>%
  PACF(daily_return, lag_max = 50) %>%
  autoplot() +
  labs(title = "PACF: Daily Returns",
       subtitle = "Values within confidence bands suggest white noise") +
  theme_minimal()

# Alternative for returns
pacf_returns <- pacf(returns_clean, lag.max = 50, plot = FALSE)
autoplot(pacf_returns) +
  labs(title = "PACF: Daily Returns (Alternative Method)",
       x = "Lag", y = "PACF") +
  theme_minimal()
```

**Key Insights:**

- **PACF of prices**: Strong lag-1 autocorrelation (AR(1) pattern)
- **PACF of returns**: Most lags within confidence bands
  - Suggests little to no autoregressive structure
  - Returns appear open to white noise
```

## Ljung-Box Test

```{r ljung-box}
# Ljung-Box test for returns (using filled gaps)
kse_returns_filled <- kse_returns %>% 
  fill_gaps() %>%
  fill(daily_return, .direction = "down")  # Fill NA with previous value

lb_test <- kse_returns_filled %>%
  features(daily_return, ljung_box, lag = 20, dof = 0)

lb_test %>%
  kable(caption = "Ljung-Box Test Results for Daily Returns",
        digits = 4,
        col.names = c("Statistic", "P-value")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Interpretation
cat("\n**Test Result:**\n\n")
if(lb_test$lb_pvalue < 0.05) {
  cat("✓ The returns show significant autocorrelation (p < 0.05)\n")
  cat("  → There is predictable structure in returns\n")
  cat("  → ARIMA models may be appropriate\n")
} else {
  cat("✗ No significant autocorrelation in returns (p ≥ 0.05)\n")
  cat("  → Returns appear to be white noise\n")
  cat("  → Market may be informationally efficient\n")
}

# Additional test on squared returns (test for ARCH effects)
kse_returns_sq <- kse_returns_filled %>%
  mutate(sq_return = daily_return^2)

lb_test_sq <- kse_returns_sq %>%
  features(sq_return, ljung_box, lag = 20, dof = 0)

cat("\n\n**Ljung-Box Test on Squared Returns:**\n\n")
lb_test_sq %>%
  kable(caption = "Testing for Volatility Clustering (ARCH Effects)",
        digits = 4,
        col.names = c("Statistic", "P-value")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

if(lb_test_sq$lb_pvalue < 0.05) {
  cat("\n✓ Significant autocorrelation in squared returns\n")
  cat("  → Volatility clustering present\n")
  cat("  → GARCH models may be appropriate\n")
}
```

# Stationarity Testing {.tabset}

## Visual Tests

```{r visual-stationarity}
# Plot with rolling mean and std
kse_rolling <- kse_tsbl %>%
  mutate(
    roll_mean = slider::slide_dbl(open, mean, .before = 29, .complete = TRUE),
    roll_sd = slider::slide_dbl(open, sd, .before = 29, .complete = TRUE)
  )

# Rolling mean
ggplot(kse_rolling, aes(x = date)) +
  geom_line(aes(y = open), alpha = 0.5, color = "gray") +
  geom_line(aes(y = roll_mean), color = "blue", size = 1) +
  labs(title = "KSE-100 with 30-Day Rolling Mean",
       y = "Price", x = "Date") +
  theme_minimal()

# Rolling standard deviation
ggplot(kse_rolling, aes(x = date, y = roll_sd)) +
  geom_line(color = "darkred", size = 1) +
  labs(title = "30-Day Rolling Standard Deviation",
       y = "Standard Deviation", x = "Date") +
  theme_minimal()
```

## Unit Root Tests

```{r unit-root}
# ADF test function
adf_test <- function(x) {
  test <- tseries::adf.test(x, alternative = "stationary")
  tibble(
    statistic = test$statistic,
    p_value = test$p.value,
    conclusion = ifelse(test$p.value < 0.05, "Stationary", "Non-Stationary")
  )
}

# KPSS test function
kpss_test <- function(x) {
  test <- tseries::kpss.test(x)
  tibble(
    statistic = test$statistic,
    p_value = test$p.value,
    conclusion = ifelse(test$p.value < 0.05, "Non-Stationary", "Stationary")
  )
}

# Test closing prices
cat("Tests for Closing Prices:\n\n")
adf_open <- adf_test(na.omit(kse$open))
kpss_open <- kpss_test(na.omit(kse$open))

bind_rows(
  adf_open %>% mutate(test = "ADF"),
  kpss_open %>% mutate(test = "KPSS")
) %>%
  select(test, everything()) %>%
  kable(caption = "Stationarity Tests: Closing Prices",
        digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Test returns
cat("\nTests for Daily Returns:\n\n")
returns_clean <- na.omit(kse_returns$daily_return)
adf_returns <- adf_test(returns_clean)
kpss_returns <- kpss_test(returns_clean)

bind_rows(
  adf_returns %>% mutate(test = "ADF"),
  kpss_returns %>% mutate(test = "KPSS")
) %>%
  select(test, everything()) %>%
  kable(caption = "Stationarity Tests: Daily Returns",
        digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Differencing

```{r differencing}
# First difference
kse_diff <- kse_tsbl %>%
  mutate(diff_open = difference(open))

# Plot original and differenced series
p1 <- kse_tsbl %>%
  autoplot(open) +
  labs(title = "Original Series", y = "open Price") +
  theme_minimal()

p2 <- kse_diff %>%
  autoplot(diff_open) +
  labs(title = "First Difference", y = "Differenced Price") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 1)

# Test differenced series
diff_clean <- na.omit(kse_diff$diff_open)
cat("Stationarity tests for differenced series:\n\n")

bind_rows(
  adf_test(diff_clean) %>% mutate(test = "ADF"),
  kpss_test(diff_clean) %>% mutate(test = "KPSS")
) %>%
  select(test, everything()) %>%
  kable(caption = "Stationarity Tests: First Difference",
        digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Simple Forecasting Models {.tabset}

## Benchmark Methods

```{r benchmark-models}
# Use last 30 days as test set
train <- kse_monthly %>% slice(1:(n()-6))
test <- kse_monthly %>% slice((n()-5):n())

# Fit benchmark models
fit_bench <- train %>%
  model(
    Mean = MEAN(avg_open),
    Naive = NAIVE(avg_open),
    Drift = RW(avg_open ~ drift()),
    SNaive = SNAIVE(avg_open)
  )

# Generate forecasts
fc_bench <- fit_bench %>% forecast(h = 6)

# Plot forecasts
fc_bench %>%
  autoplot(train, level = NULL) +
  autolayer(test, avg_open, color = "black") +
  labs(title = "Benchmark Forecasts: KSE-100 Monthly",
       y = "Average open Price",
       x = "Month") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Accuracy
accuracy(fc_bench, test) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  arrange(RMSE) %>%
  kable(caption = "Benchmark Model Accuracy",
        digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## ETS Models

```{r ets-models}
# Fit ETS model
fit_ets <- train %>%
  model(
    auto_ets = ETS(avg_open)
  )

# Report model
report(fit_ets)

# Forecast
fc_ets <- fit_ets %>% forecast(h = 6)

# Plot
fc_ets %>%
  autoplot(train) +
  autolayer(test, avg_open, color = "black") +
  labs(title = "ETS Forecast: KSE-100 Monthly",
       y = "Average open Price") +
  theme_minimal()

# Accuracy
accuracy(fc_ets, test) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  kable(caption = "ETS Model Accuracy",
        digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## ARIMA Models

```{r arima-models}
# Fit ARIMA model
fit_arima <- train %>%
  model(
    auto_arima = ARIMA(avg_open, stepwise = FALSE, approximation = FALSE)
  )

# Report model
report(fit_arima)

# Forecast
fc_arima <- fit_arima %>% forecast(h = 6)

# Plot
fc_arima %>%
  autoplot(train) +
  autolayer(test, avg_open, color = "black") +
  labs(title = "ARIMA Forecast: KSE-100 Monthly",
       y = "Average open Price") +
  theme_minimal()

# Residual diagnostics
fit_arima %>%
  gg_tsresiduals() +
  labs(title = "Residual Diagnostics: ARIMA Model")

# Accuracy
accuracy(fc_arima, test) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  kable(caption = "ARIMA Model Accuracy",
        digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Model Comparison

```{r model-comparison}
# Compare all models
all_fc <- bind_rows(
  fc_bench,
  fc_ets,
  fc_arima
)

# Plot comparison
all_fc %>%
  autoplot(train, level = NULL) +
  autolayer(test, avg_open, color = "black", size = 1.2) +
  labs(title = "Forecast Comparison: All Models",
       y = "Average open Price",
       x = "Month") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Accuracy comparison
accuracy(all_fc, test) %>%
  select(.model, RMSE, MAE, MAPE, MASE) %>%
  arrange(RMSE) %>%
  mutate(rank = row_number()) %>%
  kable(caption = "Model Accuracy Comparison (Ranked by RMSE)",
        digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(1, bold = TRUE, background = "#e8f5e9")
```

# Advanced Topics {.tabset}

## Volatility Modeling

```{r volatility}
# Calculate realized volatility
kse_vol <- kse_returns %>%
  mutate(
    sq_return = daily_return^2,
    abs_return = abs(daily_return)
  ) %>%
  mutate(
    realized_vol = sqrt(slider::slide_dbl(sq_return, mean, 
                                          .before = 19, .complete = TRUE)) * sqrt(252)
  )

# Plot volatility
kse_vol %>%
  ggplot(aes(x = date, y = realized_vol)) +
  geom_line(color = "darkred", alpha = 0.7) +
  labs(title = "Realized Volatility (20-Day, Annualized)",
       y = "Volatility (%)",
       x = "Date") +
  theme_minimal()

# Volatility clustering visualization
kse_vol %>%
  ggplot(aes(x = date, y = abs_return)) +
  geom_line(alpha = 0.5, color = "gray50") +
  geom_smooth(span = 0.1, color = "darkred", se = FALSE) +
  labs(title = "Volatility Clustering in Daily Returns",
       y = "Absolute Return (%)",
       x = "Date") +
  theme_minimal()
```

## Moving Averages

```{r moving-averages}
# Calculate multiple moving averages
kse_ma <- kse_tsbl %>%
  mutate(
    ma_20 = slider::slide_dbl(open, mean, .before = 19, .complete = TRUE),
    ma_50 = slider::slide_dbl(open, mean, .before = 49, .complete = TRUE),
    ma_200 = slider::slide_dbl(open, mean, .before = 199, .complete = TRUE)
  )

# Plot with moving averages
kse_ma %>%
  filter(date >= max(date) - years(2)) %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = open), color = "gray50", alpha = 0.5) +
  geom_line(aes(y = ma_20, color = "20-Day MA"), size = 1) +
  geom_line(aes(y = ma_50, color = "50-Day MA"), size = 1) +
  geom_line(aes(y = ma_200, color = "200-Day MA"), size = 1) +
  scale_color_manual(values = c("20-Day MA" = "blue", 
                                "50-Day MA" = "red",
                                "200-Day MA" = "darkgreen")) +
  labs(title = "KSE-100 with Moving Averages (Last 2 Years)",
       y = "open Price", x = "Date",
       color = "") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Feature Extraction

```{r features}
# Extract time series features
kse_features <- kse_monthly %>%
  features(avg_open, feature_set(pkgs = "feasts"))

# Display key features
kse_features %>%
  select(trend_strength, seasonal_strength_year, 
         acf1, pacf1, diff1_acf1) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  kable(caption = "Key Time Series Features",
        digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Spectral analysis
kse_tsbl %>%
  gg_season(open, period = "year") +
  labs(title = "Seasonal Plot by Year") +
  theme_minimal()
```

# Summary & Conclusions {.tabset}

## Key Findings

### 1. Time Series Characteristics

- **Trend**: The KSE-100 index shows [describe trend - upward/downward/cyclical]
- **Seasonality**: [Describe any seasonal patterns found]
- **Volatility**: Evidence of volatility clustering, typical of financial time series
- **Stationarity**: Price levels are non-stationary; returns are openr to stationarity

### 2. Best Performing Models

Based on our analysis:

```{r summary-table}
# Create summary table (using previous accuracy results)
tibble(
  Model = c("Benchmark (Naive)", "ETS", "ARIMA"),
  `Best For` = c("Simple baseline", "Automatic selection", "Custom tuning"),
  Strengths = c("Simple, interpretable", "Handles trend/seasonality", "Flexible specification"),
  Limitations = c("No trend capture", "Can overfit", "Requires expertise")
) %>%
  kable(caption = "Model Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### 3. Practical Insights

- **Day-of-week effects**: [Describe if found]
- **Volatility patterns**: High volatility during [specific periods]
- **Forecast horizon**: Short-term forecasts (1-7 days) more reliable than long-term

## Recommendations

### For Investors

1. **Short-term trading**: Use technical indicators (moving averages) combined with volatility measures
2. **Long-term investing**: Focus on trend components, ignore short-term noise
3. **Risk management**: Account for volatility clustering when sizing positions

### For Further Analysis

1. **Multivariate models**: Include macroeconomic variables (GDP, inflation, interest rates)
2. **Machine learning**: Explore neural networks for pattern recognition
3. **High-frequency data**: Analyze intraday patterns if available
4. **Event studies**: Impact of specific news/events on returns
5. **Regime switching**: Model different market regimes (bull/bear markets)

## Learning Resources

### Books

1. **Forecasting: Principles and Practice (3rd ed)** by Rob J. Hyndman & George Athanasopoulos
   - Free online: [otexts.com/fpp3](https://otexts.com/fpp3/)
   - The definitive guide to modern forecasting

2. **Time Series Analysis and Its Applications** by Robert Shumway & David Stoffer
   - Comprehensive coverage of time series methods

3. **Introduction to Time Series and Forecasting** by Peter Brockwell & Richard Davis
   - Mathematical foundation for time series

### Online Courses

- Coursera: "Practical Time Series Analysis"
- DataCamp: "Time Series with R"
- Rob Hyndman's YouTube channel for lectures

### R Packages

```r
# Core packages
library(fpp3)      # Forecasting bundle
library(tsibble)   # Tidy time series
library(feasts)    # Feature extraction
library(fable)     # Forecasting models

# Visualization
library(dygraphs)  # Interactive plots
library(plotly)    # Interactive graphics

# Advanced
library(prophet)   # Facebook's forecasting
library(forecast)  # Classical methods
```

## Exercises

### Exercise 1: Seasonal Analysis
Modify the seasonal decomposition to use multiplicative instead of additive. When is multiplicative decomposition more appropriate?

### Exercise 2: Model Selection
Try different ARIMA specifications manually. Compare with auto.arima(). Can you beat the automatic selection?

### Exercise 3: Forecast Evaluation
Implement rolling window cross-validation. How does model performance change across different time periods?

### Exercise 4: Trading Strategy
Design a simple moving average crossover strategy. Backtest it on the KSE-100 data. Calculate returns and maximum drawdown.

### Exercise 5: Volatility Forecasting
Implement a GARCH model for volatility forecasting. Compare with realized volatility.

## Code Repository

The complete code for this tutorial is available with:

```r
# All data processing steps
# Model fitting procedures  
# Visualization functions
# Custom helper functions
```

## References

1. Hyndman, R.J., & Athanasopoulos, G. (2021). *Forecasting: principles and practice* (3rd ed). OTexts: Melbourne, Australia.

2. Box, G. E., Jenkins, G. M., Reinsel, G. C., & Ljung, G. M. (2015). *Time series analysis: forecasting and control* (5th ed). John Wiley & Sons.

3. Shumway, R. H., & Stoffer, D. S. (2017). *Time series analysis and its applications: with R examples* (4th ed). Springer.

4. Cleveland, R. B., Cleveland, W. S., McRae, J. E., & Terpenning, I. (1990). STL: A seasonal-trend decomposition. *Journal of official statistics*, 6(1), 3-73.

# Appendix {.tabset}

## Full Code

Here's the complete workflow in a single code block:

```{r appendix-code, eval=FALSE}
# Complete KSE-100 Time Series Analysis Workflow
# ================================================

# 1. SETUP
library(tidyverse)
library(fpp3)
library(tsibble)
library(feasts)
library(forecast)
library(dygraphs)
library(xts)
library(zoo)
library(janitor)

# 2. DATA LOADING
kse_raw <- read_csv("docs/data/kse_2025.csv") %>% 
  clean_names()

# 3. DATA PREPARATION
kse <- kse_raw %>%
  mutate(
    date = as.Date(date, format = "%d-%b-%y"),
    change_percent = as.numeric(gsub("%", "", change_percent)),
    vol = as.numeric(gsub(",", "", vol))
  ) %>%
  select(date, open, high, low, close, vol, change_percent) %>%
  arrange(date) %>%
  drop_na()

kse_tsbl <- kse %>% as_tsibble(index = date)

# 4. EXPLORATORY ANALYSIS
kse_tsbl %>% autoplot(close)
kse_tsbl %>% ACF(close) %>% autoplot()
kse_tsbl %>% PACF(close) %>% autoplot()

# 5. RETURNS CALCULATION
kse_returns <- kse_tsbl %>%
  mutate(daily_return = (close - lag(close)) / lag(close) * 100)

# 6. DECOMPOSITION
kse_monthly <- kse_tsbl %>%
  mutate(year_month = yearmonth(date)) %>%
  group_by(year_month) %>%
  summarise(avg_close = mean(close)) %>%
  as_tsibble(index = year_month)

dcmp <- kse_monthly %>%
  model(stl = STL(avg_close))

components(dcmp) %>% autoplot()

# 7. MODELING
train <- kse_monthly %>% slice(1:(n()-6))
test <- kse_monthly %>% slice((n()-5):n())

fit <- train %>%
  model(
    naive = NAIVE(avg_close),
    ets = ETS(avg_close),
    arima = ARIMA(avg_close)
  )

# 8. FORECASTING
fc <- fit %>% forecast(h = 6)

# 9. EVALUATION
accuracy(fc, test)

# 10. VISUALIZATION
fc %>%
  autoplot(train) +
  autolayer(test, avg_close)
```

## Data Dictionary

Complete variable descriptions:

```{r data-dictionary}
tibble(
  Variable = c("date", "open", "high", "low", "close", "vol", "change_percent"),
  Description = c(
    "Trading date (YYYY-MM-DD format)",
    "Opening price at market open",
    "Highest price during trading day",
    "Lowest price during trading day",
    "Closing price at market close",
    "Total trading volume",
    "Daily percentage change in closing price"
  ),
  Type = c("Date", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric"),
  Unit = c("-", "Index Points", "Index Points", "Index Points", 
           "Index Points", "Shares", "Percent")
) %>%
  kable(caption = "KSE-100 Data Dictionary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Helper Functions

Custom functions used in this tutorial:

```{r helper-functions, eval=FALSE}
# Calculate multiple moving averages
calc_moving_averages <- function(data, price_col, windows = c(20, 50, 200)) {
  for(w in windows) {
    col_name <- paste0("ma_", w)
    data <- data %>%
      mutate(!!col_name := slider::slide_dbl(
        .data[[price_col]], mean, 
        .before = w-1, .complete = TRUE
      ))
  }
  return(data)
}

# Calculate volatility
calc_volatility <- function(returns, window = 20, annualize = TRUE) {
  vol <- sqrt(slider::slide_dbl(
    returns^2, mean, 
    .before = window-1, .complete = TRUE
  ))
  
  if(annualize) {
    vol <- vol * sqrt(252)  # Assuming 252 trading days
  }
  
  return(vol)
}

# Create interactive time series plot
plot_interactive_ts <- function(data, date_col, value_col, title = "") {
  xts_data <- xts(data[[value_col]], order.by = data[[date_col]])
  
  dygraph(xts_data, main = title) %>%
    dyRangeSelector() %>%
    dyOptions(colors = "#2C3E50", strokeWidth = 1.5)
}

# Rolling window cross-validation
rolling_cv <- function(data, model_fn, h = 1, window = 100) {
  n <- nrow(data)
  errors <- numeric(n - window - h + 1)
  
  for(i in seq_along(errors)) {
    train <- data[i:(i+window-1), ]
    test <- data[i+window+h-1, ]
    
    fit <- model_fn(train)
    fc <- forecast(fit, h = h)
    errors[i] <- test$value - fc$mean[h]
  }
  
  return(errors)
}

# Calculate performance metrics
calc_metrics <- function(actual, predicted) {
  errors <- actual - predicted
  
  tibble(
    RMSE = sqrt(mean(errors^2, na.rm = TRUE)),
    MAE = mean(abs(errors), na.rm = TRUE),
    MAPE = mean(abs(errors/actual) * 100, na.rm = TRUE),
    MASE = mean(abs(errors), na.rm = TRUE) / 
           mean(abs(diff(actual)), na.rm = TRUE)
  )
}
```

## Troubleshooting

Common issues and solutions:

### Issue 1: Date Parsing Errors
```r
# Problem: Dates not recognized
# Solution: Specify format explicitly
dates <- as.Date(date_column, format = "%d-%b-%y")
```

### Issue 2: Missing Values
```r
# Check for NA values
sum(is.na(data))

# Remove rows with NA
data_clean <- data %>% drop_na()

# Or interpolate
data_imputed <- data %>%
  mutate(value = zoo::na.approx(value))
```

### Issue 3: Irregular Time Series
```r
# Check for gaps
has_gaps(tsibble_data)

# Fill gaps
tsibble_data %>% fill_gaps()
```

### Issue 4: Model Convergence
```r
# ARIMA not converging
# Solution: Try different starting values or simpler models
model(ARIMA(close ~ pdq(1,1,1) + PDQ(0,0,0)))
```

### Issue 5: Memory Issues with Large Data
```r
# Use data.table for large datasets
library(data.table)
kse_dt <- fread("large_file.csv")

# Or use sampling
kse_sample <- kse %>% slice_sample(prop = 0.1)
```

## Session Info

```{r session-info}
sessionInfo()
```

---

## Contact & Support

**Questions?** Feel free to reach out:

- GitHub Issues: [your-repo/issues]
- Email: [your-email]
- Twitter: [@your-handle]

**Acknowledgments:**

This tutorial was inspired by the teaching philosophy of Professor Rob J. Hyndman and the excellent resources provided by the `fpp3` package ecosystem.

---

**Last Updated:** `r Sys.Date()`

**Version:** 1.0

**License:** MIT License - Free to use and modify with attribution

---

# Navigation Tips

- Use the **table of contents** (left sidebar) for quick navigation
- Each section has **multiple tabs** - click through them all
- **Code folding** is enabled - click "Code" to see implementation details
- All plots are **interactive** where possible - hover for details
- **Download** this document as PDF using your browser's print function

---

*"The best way to learn time series analysis is to practice with real data. The KSE-100 index provides an excellent playground for exploring concepts from trend analysis to forecasting."*

**Happy Forecasting! 📈**