---
title: "Spurious Regression"
title-block-banner: true
author: "Zahid Asghar"
#date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    chalkboard: true
    #   buttons: false
    preview-links: auto
    title-slide-attributes:
      data-background-color: "#cf3a36"
overview: true
execute:
  echo: true
  warning: false
  freeze: auto
---

# Spurious Regression

$y_t = a_0 + a_1z_t + e_t$ 

where $y_t$ is the dependent variable, $z_t$ is the independent variable, $a_0$ and $a_1$ are the parameters, and $e_t$ is the error term.
Assumptions of the classical model: 

-   both the {$y_t$} and {$z_t$} sequences be stationary 

-   the errors have a zero mean and a finite variance. 

## 

In the presence of nonstationary variables, there might be what Granger and Newbold (1974) call a spurious regression.  

> A spurious regression has a high $R^2$ and $t$-statistics that appear to be significant, but the results are without any economic meaning. The regression output “looks good” because the least-squares estimates are not consistent and the customary tests of statistical inference do not hold.

## {.fragment} 

--- 

**CASE 1:** Both  {yt} and {zt} are stationary. The classical regression model is appropriate.


**CASE 2:** The {yt} and {zt} sequences are integrated of different orders. Regression equations using such variables are meaningless. 

**CASE 3:** The nonstationary {yt} and {zt} sequences are integrated of the same order and the residual sequence contains a stochastic trend. This is the case in which the regression is spurious. In this case, it is often recommended that the regression equation be estimated in first differences. 
---
**CASE 4:** The nonstationary {yt} and {zt} sequences are integrated of the same order and the residual sequence is stationary. In this circumstance, {yt} and {zt} are cointegrated.

## Load libraries and generate data


```{r}
library(tidyverse) # An umbrella package that installs the tidyverse packages

library(readxl) # To read excel files

library(broom) # To tidy up the regression output

library(moderndive) # To get regression points

library(ggfortify) # To plot regression diagnostics

library(ggResidpanel) # To plot regression diagnostics
```


```{r}
T <- 1000
set.seed(1357)
y <- ts(rep(0,T))
vy <- ts(rnorm(T))
for (t in 2:T){
  y[t] <- y[t-1]+vy[t]
}

set.seed(4365)
x <- ts(rep(0,T))
vx <- ts(rnorm(T))
for (t in 2:T){
  x[t] <- x[t-1]+vx[t]
}
y <- ts(y[300:1000])
x <- ts(x[300:1000])
ts.plot(y,x, ylab="y and x")

``` 




## 

```{r}


# barplot of residuals
resid <- residuals(lm(y~x))
barplot(resid, ylab="Residuals", xlab="Observation", main="Residuals of y~x")

```

## Residuals panel

```{r}
resid_panel(lm(y~x))
```

