---
title: "VAR and SVAR in R"
subtitle: "Vector Autoregression by Stock and Watson JEP-2001"
author: "Zahid Asghar"
date: 12/14/2023
date-format: long
format: 
  html:
    toc: true
    toc_float: true
    number_sections: true
    theme: solarized
    highlight: tango
    code_folding: true
execute:
  freeze: auto
  warning: false
---

# Vector Autoregression

## Introduction

In this post I am going to reproduce results of article by [**Stock, James H., and Mark W. Watson, "Vector Autoregressions," Journal of Economic Perspectives, Vol. 15 No. 4 (Fall 2001), 101-115.**](https://www.aeaweb.org/articles?id=10.1257/jep.15.4.101) using R.

## Load necessary libraries

```{r}
#| warnings: false
library(tidyverse) # An umbrella library consisting of 9 R packages
library(vars)
library(svars)
library(lmtest)
library(forecast)
library(haven) # To read data from stata/spss
```

## Load data

```{r}
sw_df <- read_dta("C:/Users/92300/Dropbox/Applied Econometrics SBP/VAR JEP data.dta")
```

## Inspect data

`head()` and `tail()` provides first 6 and last 6 observations, respectively. One can use `glimpse` to see nature of each variable.

```{r}
sw_df |> head()
sw_df |> tail()
```

## Set the time series data

```{r}
sw_df <- ts(sw_df, start = c(1960, 1), frequency = 4)

sw_df1 <- sw_df[,2:4]

## Plot the time series data
plot(sw_df1, main = "Time Series Plot", ylab = "Percent", xlab = "Year")
```

## Estimate Reduced form model

As I have explained ... video that reduced VAR is used only for forecasting and it has lags of all endogenous variables. First we select appropriate lags using `lag length selection criteria` as follows:

```{r}
# Estimate VAR model
VARselect(sw_df1, lag.max = 4,
          type = "const", season = 4)
```

## Estimation of VAR model with lag order 4

In original article there are four lags selected and AIC and FPE is also indicating 4 lags, therefore, I estimate VAR model with 4 lags:

```{r}
var_model <- VAR(sw_df1, p = 4, type = "const")
summary(var_model)
```

::: callout-note
Individual coefficients in VAR model should not be interpretted and current VAR model form is only for forecasting.
:::

## Perform granger causality test

Assuming var_model is your VAR model with lag order 4

```{r}
# Perform Granger causality test for each effect variable separately
granger_causality_unrate <- causality(var_model, cause = "inflation")
granger_causality_ffr <- causality(var_model, cause = "unrate")
granger_causality_inf <- causality(var_model, cause = "ffr")
granger_causality_inf
granger_causality_unrate
granger_causality_ffr
```

## If one has to go by pair of variables

::: callout-caution
Granger causality is test of descriptive statistics, so dont interpret it in structural inference sense. For reference, please read paper I am reproducing here.
:::

```{r}

grangertest(inflation ~ ffr,order = 4, data = sw_df1)

grangertest(inflation ~ unrate,order = 4, data = sw_df1)

```

## Forecast Error Variance Decomposition

Another descriptive measure in reduced form VAR model

Forecast error variance decomposition only at 4, 8 and 12 quarters

```{r}
fevd_model1<- fevd(var_model, n.ahead = 1)
fevd_model2 <- fevd(var_model, n.ahead = 4)
fevd_model3 <- fevd(var_model, n.ahead = 8)

fevd_model4 <- fevd(var_model, n.ahead = 12)

fevd_model1

```

## Impulse response function

Third measure of descriptive statistics for reduced VAR. IRF is also from SVAR, so its important to distinguish which VAR model one is working with.

```{r}
# Estimate impulse response function from VAR
irf_model <- irf(var_model, impulse = c("inflation", "unrate", "ffr"), response = c("inflation", "unrate", "ffr"), n.ahead = 20)

plot(irf_model, yline = 0, ncol = 3, by = "response", main = "Impulse Response Function")
```

## Structural VAR

Difference between Reduced VAR, Recursive VAR and SVAR is already explained in ...video. Therefore, I am only estimating it here using `R`.

```{r}
# Define the A and B matrices for SVAR
A <- matrix(c(1, 0, 0, NA, 1, 0, NA, NA, 1), nrow = 3, byrow = TRUE)
B <- matrix(c(NA, 0, 0, 0, NA, 0, 0, 0, 1), nrow = 3, byrow = TRUE)
```

```{r}
# Perform SVAR
svar_model <- SVAR(var_model, A = A, B = B)

svar_model
```

```{r}

irf_ffr <- irf(svar_model, impulse = "ffr", response = c("inflation", "unrate", "ffr"), boot =
      FALSE)

plot(irf_ffr)
```
